---
title: "cogmod"
format: gfm
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.path: "man/figures/"
---

Cognitive Models for Subjective Scales and Decision Making Tasks in R

## Installation

```{r}
#| eval: false

if(!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")

remotes::install_github("DominiqueMakowski/cogmod")
```

## CHOCO Model

The Choice-Confidence (CHOCO) model is useful to model data from subjective scales, such as Likert-type or analog scales, in which the left and the right side correspond to different processes or higher order categorical responses (e.g., "disagree" vs. "agree", "true" vs. "false"). They can be used to jointly model choice (left or right) and confidence (the degree of left or right).

```{r}
#| fig-width: 10
#| fig-height: 7
#| code-fold: true

library(ggplot2)
library(cogmod)

# Simulate data using rchoco() with two parameter sets
df1 <- rchoco(n = 5000, mu = 0.5, muleft = 0.4, phileft = 3, kleft = 0.9)
df2 <- rchoco(n = 5000, mu = 0.7, muleft = 0.6, phileft = 5, kleft = 0.95)

# Combine data into a single data frame
df <- data.frame(
  value = c(df1, df2),
  group = rep(c("mu=0.5, muleft=0.4, phileft=3, kleft=0.9", 
                "mu=0.7, muleft=0.6, phileft=5, kleft=0.95"), each =5000)
)

# Create the histogram
ggplot(df, aes(x = value, fill = group)) +
  geom_histogram(alpha = 0.8, position = "identity", bins = 50) +
  labs(title = "CHOCO Distribution", x = "Value", y = "", fill = "Parameters") +
  theme_minimal() + 
  scale_fill_manual(values = c("#9C27B0", "#FF9800"))
```
