---
title: "cogmod"
format: gfm
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.path: "man/figures/"
editor_options: 
  chunk_output_type: console
---


[![Documentation](https://img.shields.io/badge/documentation-cogmod-orange.svg?colorB=E91E63)](https://dominiquemakowski.github.io/cogmod/)
[![Models](https://img.shields.io/badge/models-list-orange.svg?colorB=2196F3)](https://dominiquemakowski.github.io/cogmod/reference/index.html)

*Cognitive Models for Subjective Scales and Decision Making Tasks in R*

## Status

![Status](https://img.shields.io/badge/status-WIP-orange.svg)

**This package is very much totally exploratory - currently made for my own needs.** It's not meant to be stable and robust at this stage. Use at your own risks. 

- If you have suggestions for improvement, please get in touch!
- I've been seeking the best way to implement various sequential models for a long time, initially trying and [failing in R](https://github.com/DominiqueMakowski/easyRT), then developing a lot of hopes for a Julia solution - but that's not there *yet*, so I'm back at making some new attempts in R.
  - If you are interested in Sequential Sampling Models, see this amazing [Julia package](https://github.com/itsdfish/SequentialSamplingModels.jl)
  - See also this attempt at [**creating tutorials**](https://dominiquemakowski.github.io/CognitiveModels/)
  
## Installation

```{r}
#| eval: false

if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")

remotes::install_github("DominiqueMakowski/cogmod")
```

## Available Distributions 


### CHOCO Model

The [**Choice-Confidence (CHOCO)**](https://dominiquemakowski.github.io/cogmod/reference/rchoco.html) model is useful to model data from subjective ratings, such as Likert-type or analog scales, in which the left and the right side correspond to different processes or higher order categorical responses (e.g., "disagree" vs. "agree", "true" vs. "false"). They can be used to jointly model choice (left or right) and confidence (the degree of left or right).

```{r}
#| fig-width: 10
#| fig-height: 7
#| code-fold: true
#| message: false
#| warning: false

library(ggplot2)
library(patchwork)
library(cogmod)

# Simulate data using rchoco() with two parameter sets
df1 <- rchoco(n = 5000, mu = 0.5, muleft = 0.4, phileft = 3, pex = 0.1)
df2 <- rchoco(n = 5000, mu = 0.7, muleft = 0.6, phileft = 5, pex = 0.05)

# Combine data into a single data frame
df <- data.frame(
  value = c(df1, df2),
  group = rep(c(
    "mu = 0.5, muleft = 0.4, phileft = 3, pex = 0.1",
    "mu = 0.7, muleft = 0.6, phileft = 5, pex = 0.05"
  ), each = 5000)
)

# Create the histogram
ggplot(df, aes(x = value, fill = group)) +
  geom_histogram(alpha = 0.8, position = "identity", bins = 50) +
  labs(title = "CHOCO Distribution", x = "Value", y = "", fill = "Parameters") +
  theme_minimal() +
  scale_fill_manual(values = c("#9C27B0", "#FF9800"))
```

### LNR Model

The Log-Normal Race (LNR) model is useful for modeling reaction times and errors in decision-making tasks. The model assumes that each accumulator draws a value from a LogNormal distribution (shifted by a non-decision time τ). The winning accumulator (minimum draw) determines the observed reaction time and choice.

```{r}
#| fig-width: 10
#| fig-height: 7
#| code-fold: true

# Simulate data using rlnr()
lnr_data <- rlnr(n = 5000, mu = 1, mudelta = 0.5, sigmazero = 1, sigmadelta = -0.5, ndt = 0.2)

# Create histograms for each choice
ggplot(lnr_data, aes(x = rt, fill = factor(response))) +
  geom_histogram(alpha = 0.8, position = "identity", bins = 50) +
  labs(title = "LogNormal Race Model", x = "Reaction Time", y = "Frequency", fill = "Choice") +
  theme_minimal() +
  scale_fill_manual(values = c("#4CAF50", "#FF5722"))
```


## Usage with `brms`


### Subjective Ratings


#### Simulate Data

```{r}
#| fig-width: 10
#| fig-height: 7
#| warning: false
#| message: false

options(mc.cores = parallel::detectCores() - 2)

library(easystats)
library(brms)
library(cmdstanr)

df <- data.frame()
for(x in seq(0.1, 1, by = 0.1)) {
  df <- data.frame(x = x,
                   score = rchoco(n = 100, p = 0.4 + x / 2, muleft = 0.3 + x / 3, 
                                  phileft = 3, pex = 0.03, bex = 0.6, pmid = 0)) |> 
    rbind(df)
}

df |>
  ggplot(aes(x = score, y = after_stat(density))) +
  geom_histogram(bins = 100, fill = "#2196F3") +
  labs(title = "Rating Distribution", x = "Score", y = "Density") +
  theme_minimal()
```


#### ZOIB Model

The Zero-One Inflated Beta (ZOIB) model assumes that the data can be modeled as a mixture of two logistic regression processes for the boundary values (0 and 1) and a beta regression process for the continuous proportions in-between.


```{r}
#| eval: false

f <- bf(
  score ~ x,
  phi ~ x,
  zoi ~ x,
  coi ~ x
)

m_zoib <- brm(f,
  data = df, family = zero_one_inflated_beta(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_zoib <- brms::add_criterion(m_zoib, "loo")  # For later model comparison

saveRDS(m_zoib, file = "man/figures/m_zoib.rds")
```


#### XBX Model

[Kosmidis & Zeileis (2024)](https://arxiv.org/abs/2409.07233) introduce a generalization of the classic beta regression model with extended support [0, 1]. Specifically, the extended-support beta distribution (`xbeta`) leverages an underlying symmetric four-parameter beta distribution with exceedence parameter nu to obtain support [-nu, 1 + nu] that is subsequently censored to [0, 1] in order to obtain point masses at the boundary values 0 and 1.

```{r}
#| eval: false

f <- bf(
  score ~ x,
  phi ~ x,
  kappa ~ x
)

m_xbx <- brm(f,
  data = df, family = xbeta(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_xbx <- brms::add_criterion(m_xbx, "loo")  # For later model comparison

saveRDS(m_xbx, file = "man/figures/m_xbx.rds")
```


#### BEXT Model

The [**BeXt model**](https://dominiquemakowski.github.io/cogmod/reference/rbext.html) corresponds to a reparametrized ordered beta model ([Kubinec, 2023](https://doi.org/10.1017/pan.2022.20)).
Instead of defining left and right cutpoints, the BeXt parametrization
uses the likelihood of extreme values (0 and 1) and their balance (i.e., the
relative proportion of zeros and ones).

```{r}
#| eval: false

f <- bf(
  score ~ x,
  phi ~ x,
  pex ~ x,
  bex ~ x
)

m_bext <- brm(f,
  data = df, family = bext(), stanvars = bext_stanvars(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_bext <- brms::add_criterion(m_bext, "loo")  # For later model comparison

saveRDS(m_bext, file = "man/figures/m_bext.rds")
```

#### CHOCO Model

See the [**documentation**](https://dominiquemakowski.github.io/cogmod/reference/rchoco.html) of the Choice-Confidence (CHOCO).

```{r}
#| eval: false

f <- bf(
  score ~ x,
  muleft ~ x,
  mudelta ~ x,
  phileft ~ x,
  phidelta ~ x,
  pex ~ x,
  bex ~ x,
  pmid = 0
)

m_choco <- brm(f,
  data = df, family = choco(), stanvars = choco_stanvars(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_choco <- brms::add_criterion(m_choco, "loo")  # For later model comparison

saveRDS(m_choco, file = "man/figures/m_choco.rds")
```


#### Model Comparison

We can compare these models together using the `loo` package, which shows
that CHOCO provides a significantly better fit than the other models.

```{r}
#| code-fold: true

m_zoib <- readRDS("man/figures/m_zoib.rds")
m_xbx <- readRDS("man/figures/m_xbx.rds")
m_bext <- readRDS("man/figures/m_bext.rds")
m_choco <- readRDS("man/figures/m_choco.rds")

loo::loo_compare(m_zoib, m_xbx, m_bext, m_choco) |> 
  parameters(include_ENP = TRUE)
```

Running posterior predictive checks allows to visualize the predicted distributions from various models.
We can see how typical Beta-related models fail to capture the bimodal nature of the data, which is 
well captured by the CHOCO model.


```{r}
#| code-fold: true

pred <- rbind(
  estimate_prediction(m_zoib, keep_iterations = 200) |>
    reshape_iterations() |>
    data_modify(Model = "ZOIB"),
  estimate_prediction(m_xbx, keep_iterations = 200) |>
    reshape_iterations() |>
    data_modify(Model = "XBX"),
  estimate_prediction(m_bext, keep_iterations = 200) |>
    reshape_iterations() |>
    data_modify(Model = "BEXT"),
  estimate_prediction(m_choco, keep_iterations = 200) |>
    reshape_iterations() |>
    data_modify(Model = "CHOCO")
)

insight::get_data(m_zoib) |>
  ggplot(aes(x = score, y = after_stat(density))) +
  geom_histogram(bins = 100, fill = "#2196F3") +
  labs(title = "Rating Distribution", x = "Score", y = "Density") +
  theme_minimal() + 
  geom_histogram(
    data = pred, aes(x = iter_value, group = as.factor(iter_group)),
    bins = 100, alpha = 0.02, position = "identity", fill = "#FF5722"
  ) +
  facet_wrap(~Model)
```


#### Effect Visualisation

We can see how the predicted distribution changes as a function of **x** and gets "pushed" to the right. 
Moreover, we can also visualize the effect of **x** on specific parameters, showing that it mostly affects the main parameters **mu**, 
which corresponds to the ***p*** probability of answering on the right. This is consistent with our expectations, and reflects 
the larger mass on the right of the scale for higher value of **x** (in brown).

```{r}
#| fig-width: 10
#| fig-height: 8
#| code-fold: true

p1 <- modelbased::estimate_prediction(m_choco, data = "grid", length = 4, keep_iterations = 500) |> 
  reshape_iterations() |> 
  ggplot(aes(x = iter_value, fill = as.factor(x))) +
  geom_histogram(alpha = 0.6, bins = 100, position = "identity") +
  scale_fill_bluebrown_d() +
  theme_minimal()

# Predict various parameters
pred_params <- data.frame()
for(param in c("mu", "pex", "phileft", "bex")) {
  pred_params <- m_choco |> 
    modelbased::estimate_prediction(data = "grid", length = 20, predict = param) |>
    as.data.frame() |> 
    dplyr::mutate(Parameter = param) |>  # TODO: replace by data_modify after PR
    rbind(pred_params)
}

p2 <- pred_params |> 
  ggplot(aes(x = x, y = Predicted)) +
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high, fill = Parameter), alpha = 0.2) +
  geom_line(aes(color = Parameter), linewidth = 1) +
  facet_wrap(~Parameter, scales = "free_y", ncol=4) +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  theme_minimal()

p1 / p2
```



### Decision Making (Choice + RT)


#### Simulate Data

```{r}
#| fig-width: 10
#| fig-height: 7
#| warning: false
#| message: false

df <- rlnr(n = 2000, mu = 0, mudelta = 0.3, sigmazero = 1, sigmadelta = -0.5, ndt = 0.2) |> 
  datawizard::data_filter(rt < 5)

df |>
  ggplot(aes(x = rt, fill = factor(response))) +
  geom_histogram(alpha = 0.8, position = "identity", bins = 100) +
  labs(title = "RT Distribution", x = "Reaction Time", y = "Frequency", fill = "Choice") +
  theme_minimal() +
  scale_fill_manual(values = c("#009688", "#E91E63"))
```

#### Drift Diffusion Model (DDM)

```{r}
#| eval: false

options(mc.cores = parallel::detectCores() - 2)

f <- bf(
  rt | dec(response) ~ 1,
  bs ~ 1,
  bias ~ 1,
  ndt ~ 1
)

m_ddm <- brm(f,
  data = df, family = wiener(), 
  init = \() list(Intercept = 0, Intercept_bs = 0, Intercept_bias = 0, Intercept_ndt = log(0.2)),
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_ddm <- brms::add_criterion(m_ddm, "loo") 

saveRDS(m_ddm, file = "man/figures/m_ddm.rds")
```

```{r}
#| eval: false

m_ddm <- readRDS("man/figures/m_ddm.rds")

# parameters::parameters(m_ddm, component = "all")

# library(brms)
# library(cmdstanr)
#
# m_ddm <- readRDS(url("https://raw.github.com/DominiqueMakowski/cogmod/main/man/figures/m_ddm.rds"))
# # m_ddm
#
# insight::get_predicted(m_ddm, iterations = 5, data = insight::get_data(m_ddm)[1:4,]) |>
#   as.data.frame()
#
# brms::posterior_predict(m_ddm, ndraws=5, newdata = insight::get_data(m_ddm)[1:4,], negative_rt = TRUE)
```

#### LogNormal Race (LNR)

```{r}
#| eval: false

f <- bf(
  rt | dec(response) ~ 1,
  mudelta ~ 1,
  sigmazero ~ 1,
  sigmadelta ~ 1,
  tau ~ 1,
  minrt = min(df$rt)
)

m_lnr <- brm(f,
  data = df,
  init = 1,
  family = lnr(),
  stanvars = lnr_stanvars(),
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_lnr <- brms::add_criterion(m_lnr, "loo")

saveRDS(m_lnr, file = "man/figures/m_lnr.rds")
```

```{r}
#| eval: false

# m_lnr <- readRDS("man/figures/m_lnr.rds")
# 
# parameters::parameters(m_lnr, component = "all")

# library(brms)
# library(cmdstanr)
#
# m_ddm <- readRDS(url("https://raw.github.com/DominiqueMakowski/cogmod/main/man/figures/m_ddm.rds"))
# # m_ddm
#
# insight::get_predicted(m_ddm, iterations = 5, data = insight::get_data(m_ddm)[1:4,]) |>
#   as.data.frame()
#
# brms::posterior_predict(m_ddm, ndraws=5, newdata = insight::get_data(m_ddm)[1:4,], negative_rt = TRUE)

# d <- data.frame(rt = runif(100), response = sample(c(0, 1), 100, replace = TRUE))
# brms::make_stancode(rt ~ 1, family = brms::shifted_lognormal(), data = d)
# brms::make_stancode(brms::bf(
#   rt | dec(response) ~ 1,
#   bs ~ 1,
#   bias ~ 1,
#   ndt ~ 1
# ), family = brms::wiener(), data = d)
# brms::make_stancode(brms::bf(
#   rt | dec(response) ~ 1,
#   mudelta ~ 1,
#   sigmazero ~ 1,
#   sigmadelta ~ 1,
#   tau ~ 1
# ), family = lnr(), stanvar = lnr_stanvars(df$rt), data = df)
```

#### Model Comparison

```{r}
m_ddm <- readRDS("man/figures/m_ddm.rds")
m_lnr <- readRDS("man/figures/m_lnr.rds")

loo::loo_compare(m_ddm, m_lnr) |> 
  parameters::parameters()
```

