---
title: "cogmod"
format: gfm
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.path: "man/figures/"
editor_options: 
  chunk_output_type: console
---

Cognitive Models for Subjective Scales and Decision Making Tasks in R

## Status

**This package is very much totally exploratory - currently made for my own needs.** It's not meant to be stable and robust at this stage. Use at your own risks. 

- If you have suggestions for improvement, please get in touch!
- I've been seeking the best way to implement various sequential models for a long time, initially trying and [failing in R](https://github.com/DominiqueMakowski/easyRT), then developing a lot of hopes for a Julia solution - but that's not there *yet*, so I'm back at making some new attempts in R.
  - If you are interested in Sequential Sampling Models, see this amazing [Julia package](https://github.com/itsdfish/SequentialSamplingModels.jl)
  - See also this attempt at [**creating tutorials**](https://dominiquemakowski.github.io/CognitiveModels/)
  
## Installation

```{r}
#| eval: false

if(!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")

remotes::install_github("DominiqueMakowski/cogmod")
```

## CHOCO Model

The Choice-Confidence (CHOCO) model is useful to model data from subjective scales, such as Likert-type or analog scales, in which the left and the right side correspond to different processes or higher order categorical responses (e.g., "disagree" vs. "agree", "true" vs. "false"). They can be used to jointly model choice (left or right) and confidence (the degree of left or right).

```{r}
#| fig-width: 10
#| fig-height: 7
#| code-fold: true

library(ggplot2)
library(cogmod)

# Simulate data using rchoco() with two parameter sets
df1 <- rchoco(n = 5000, mu = 0.5, muleft = 0.4, phileft = 3, kleft = 0.9)
df2 <- rchoco(n = 5000, mu = 0.7, muleft = 0.6, phileft = 5, kleft = 0.95)

# Combine data into a single data frame
df <- data.frame(
  value = c(df1, df2),
  group = rep(c("mu=0.5, muleft=0.4, phileft=3, kleft=0.9", 
                "mu=0.7, muleft=0.6, phileft=5, kleft=0.95"), each =5000)
)

# Create the histogram
ggplot(df, aes(x = value, fill = group)) +
  geom_histogram(alpha = 0.8, position = "identity", bins = 50) +
  labs(title = "CHOCO Distribution", x = "Value", y = "", fill = "Parameters") +
  theme_minimal() + 
  scale_fill_manual(values = c("#9C27B0", "#FF9800"))
```

## LNR Model

The Log-Normal Race (LNR) model is useful for modeling reaction times and errors in decision-making tasks. The model assumes that each accumulator draws a value from a LogNormal distribution (shifted by a non-decision time τ). The winning accumulator (minimum draw) determines the observed reaction time and choice.

```{r}
#| fig-width: 10
#| fig-height: 7
#| code-fold: true

# Simulate data using rlnr()
lnr_data <- rlnr(n = 5000, mu = 1, mud = 0.5, sigmazero = 1, sigmad = -0.5, tau = 0.2)

# Create histograms for each choice
ggplot(lnr_data, aes(x = rt, fill = factor(choice))) +
  geom_histogram(alpha = 0.8, position = "identity", bins = 50) +
  labs(title = "LNR Distribution", x = "Reaction Time", y = "Frequency", fill = "Choice") +
  theme_minimal() +
  scale_fill_manual(values = c("#4CAF50", "#FF5722"))
```


## Usage with `brms`

### Simulate Data

```{r}
#| fig-width: 10
#| fig-height: 7
#| warning: false
#| message: false

options(mc.cores = parallel::detectCores() - 2)

library(brms)
library(cmdstanr)

df <- brms::rwiener(n = 5000, delta = 0.5, alpha = 1, beta = .3, tau = .25) |> 
  datawizard::data_rename(replacement = c("rt", "choice")) |> 
  datawizard::data_filter(rt < 2)

df |> 
  ggplot(aes(x = rt, fill = factor(choice))) +
  geom_histogram(alpha = 0.8, position = "identity", bins = 100) +
  labs(title = "RT Distribution", x = "Reaction Time", y = "Frequency", fill = "Choice") +
  theme_minimal() +
  scale_fill_manual(values = c("#009688", "#E91E63"))
```

### Drift Diffusion Model (DDM)

```{r}
#| eval: false

f <- bf(rt | dec(choice) ~ 1,
        bs ~ 1,
        bias ~ 1,
        ndt ~ 1)

m_ddm <- brm(f, data = df, family = wiener(link = "identity", link_bs = "log", link_bias = "logit", link_ndt = "log"), 
             chains=4, iter = 500, backend="cmdstanr")

saveRDS(m_ddm, file = "man/figures/m_ddm.rds")
```

```{r}
#| eval: false

m_ddm <- readRDS("man/figures/m_ddm.rds")

parameters::parameters(m_ddm, component = "all")

# insight::get_predicted(m_ddm)
brms::posterior_predict(m_ddm, ndraws=5, newdata = df[1:4, ], negative_rt = TRUE)
```




