---
title: "Decision Making Models"
vignette: >
  %\VignetteIndexEntry{Decision Making Models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
format: 
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(cogmod)
library(easystats)
library(ggplot2)
library(brms)
library(cmdstanr)

options(mc.cores = parallel::detectCores() - 2)
```

## The Data

Decision making models jointly account for the **choice** that was made and
the **response time (RT)** it took to make it. Rather than simulating data,
we re-use the `wagenmakers2008` dataset (see the
[RT-only Models](rt_models.html) vignette) - but this time, instead of
discarding the errors, we model **choice** as *correct* vs. *error* responses.
This is a common strategy in the decision-making literature when the task
itself does not have a natural "left vs. right" stimulus category to map onto
the two accumulators/boundaries of the models below.

Evidence accumulation models are considerably more
expensive to sample than the RT-only models. We therefore use a smaller
subset of the data here, only one participant, so that
the models below can be fit in a reasonable amount of time for demonstration purposes.

```{r}
#| code-fold: false

set.seed(123)  # For reproducibility

df <- cogmod::wagenmakers2008
df <- df[df$Participant == 1, ]

# Show 10 first rows
head(df[c("Participant", "Condition", "RT", "Error")], 10)
```

```{r}
ggplot(df, aes(x = RT, fill = factor(Error))) +
  geom_histogram(bins = 100, alpha = 0.8, position = "identity") +
  facet_wrap(~Condition) +
  labs(x = "RT (s)", fill = "Response") +
  scale_fill_manual(values = c("darkgreen", "darkred"), labels = c("Correct", "Error")) +
  theme_minimal()
```

Errors are much rarer than correct responses (especially in the `Accuracy`
condition), which can be problematic for accurate estimations.

## Models

All four models below use `dec(Error)` to indicate the two-choice
outcome (`0` = Correct, `1` = Error), and a fixed `minrt` (the minimum
observed RT) to scale the non-decision time parameter `tau`.

### Drift Diffusion Model (DDM)

The DDM assumes that evidence accumulates towards one of two boundaries at
a rate `mu` (drift rate). `bs` is the boundary separation (higher = more
cautious), `bias` is the starting point between the two boundaries
(`0.5` = unbiased), and `tau` is the non-decision time (as a proportion of
`minrt`).

Note that we use the "simple" 4-parameter DDM here, which does not include between-trial variability in drift rate, starting point, or non-decision time, hence the `sigmadrift`, `sigmabias`, and `sigmatau` parameters are fixed to `0`. Estimating these parameters is possible, but considerably more expensive and often unnecessary for many applications.

```{r}
#| eval: false
#| code-fold: false

f <- bf(
  RT | dec(Error) ~ Condition,
  bs ~ Condition,
  bias ~ 1,
  tau ~ 1,
  minrt = min(df$RT),
  sigmadrift = 0,
  sigmabias = 0,
  sigmatau = 0,
  family = ddm()
)

m_ddm <- brm(f,
  data = df,
  init = 0,
  stanvars = ddm_stanvars(),
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_ddm <- brms::add_criterion(m_ddm, "loo")  # Add model performance criterion
```
```{r}
#| eval: false
#| echo: false

saveRDS(m_ddm, file = "models/m_ddm.rds")
```


### LogNormal Race (LNR)

The LNR is a somewhat simpler model. It is similar to the LBA, but each accumulator's finishing time is
drawn directly from a LogNormal distribution instead of a ballistic
accumulation process. `mu` (`nuzero`) and `nuone` are the (inverse
log-space mean) processing speeds for the "Error" and "Correct"
accumulators, and `sigmazero`/`sigmaone` their log-space SDs.

```{r}
#| eval: false
#| code-fold: false

f <- bf(
  RT | dec(Error) ~ Condition,
  nuone ~ Condition,
  sigmazero ~ 1,
  sigmaone ~ 1,
  tau ~ 1,
  minrt = min(df$RT),
  family = lnr()
)

priors <- brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau") |>
  brms::validate_prior(f, data = df)

m_lnr <- brm(f,
  data = df,
  # prior = priors,
  init = 0,
  stanvars = lnr_stanvars(),
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_lnr <- brms::add_criterion(m_lnr, "loo")  # Add model performance criterion
```
```{r}
#| eval: false
#| echo: false

saveRDS(m_lnr, file = "models/m_lnr.rds")
```



### Linear Ballistic Accumulator (LBA)

The LBA assumes two independent accumulators (one per choice) that race
towards a common threshold `b` (`sigmabias` = start-point range `A`, `bs` =
extra distance so that `b = A + bs`). `mu` and `driftone` are the mean
drift rates for the "Correct" and "Error" accumulators, and `sigmazero`/
`sigmaone` their between-trial drift variability.

```{r}
#| eval: false
#| code-fold: false

f <- bf(
  RT | dec(Error) ~ Condition,
  driftone ~ Condition,
  sigmazero ~ 1,
  sigmaone ~ 1,
  sigmabias ~ 1,
  bs ~ 1,
  tau ~ 1,
  minrt = min(df$RT),
  family = lba()
)

priors <- c(
  brms::set_prior("normal(0, 2)", class = "Intercept"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "driftone"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "sigmazero"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "sigmaone"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "sigmabias"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "bs"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau")
) |>
  brms::validate_prior(f, data = df)

m_lba <- brm(f,
  data = df,
  prior = priors,
  init = 0.5,
  stanvars = lba_stanvars(),
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_lba <- brms::add_criterion(m_lba, "loo")  # Add model performance criterion
```
```{r}
#| eval: false
#| echo: false

saveRDS(m_lba, file = "models/m_lba2.rds")
```


### Racing Diffusion Model (RDM)

The RDM is the LBA's stochastic counterpart - each accumulator integrates
evidence through the DDM's random walk process instead of accumulating
linearly. It otherwise keeps the racing architecture: two independent
accumulators, a common threshold `b`, and a start point drawn from
`Uniform(0, sigmabias)`. The consequence of swapping the ballistic path
for a diffusing one is that the noise now lives *within* a trial rather
than between trials. That is the point of the
model: Tillman et al. ([2020](https://doi.org/10.3758/s13423-020-01738-8))
show that within-trial variability alone accounts for the benchmark
choice-RT phenomena, without the between-trial drift variability that the
LBA (`sigmazero`/`sigmaone`) and the full DDM (`sigmadrift`) need. It is
therefore the more parsimonious race: `mu` and `driftone` are the drift
rates for the "Correct" and "Error" accumulators, and there is no drift
variability parameter to estimate.

Because each accumulator is a Wald (shifted inverse Gaussian) process, the
drift rates use a softplus link and are constrained to be non-negative.

```{r}
#| eval: false
#| code-fold: false

f <- bf(
  RT | dec(Error) ~ Condition,
  driftone ~ Condition,
  sigmabias ~ 1,
  bs ~ 1,
  tau ~ 1,
  minrt = min(df$RT),
  family = rdm()
)

priors <- c(
  brms::set_prior("normal(0, 2)", class = "Intercept"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "driftone"),
  # See the note below: sigmabias needs a prior to stay identified.
  brms::set_prior("normal(-1, 1)", class = "Intercept", dpar = "sigmabias"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "bs"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau")
) |>
  brms::validate_prior(f, data = df)

m_rdm <- brm(f,
  data = df,
  prior = priors,
  init = 0.5,
  stanvars = rdm_stanvars(),
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_rdm <- brms::add_criterion(m_rdm, "loo")  # Add model performance criterion
```
```{r}
#| eval: false
#| echo: false

saveRDS(m_rdm, file = "models/m_rdm.rds")
```

### Identifiability of `sigmabias` and `bs`

This concerns the LBA as much as the RDM, since both are parameterized the
same way. The two parameters enter the model only through the threshold
`b = bs + sigmabias`, so they trade off almost freely: on simulated data
with 4,000 trials, the profile log-likelihood moves by about 3 units as
`sigmabias` sweeps from 0 to half the threshold, while `bs` slides to
compensate. Under flat priors the sampler drifts down the
`sigmabias -> 0` ridge (which is just a race of two plain Wald
accumulators) and produces divergent transitions - several hundred of
them, in a 4,000-trial simulation.

A weakly informative prior on `sigmabias` is enough to prevent that: under
the softplus link, `normal(-1, 1)` is centred near `0.31`, and it removes
the divergences entirely. It does not make `sigmabias` itself trustworthy,
though. In a recovery check where the true value was `0.3` and the prior
was centred at `0.31`, the posterior still came back at `0.21` - the
likelihood pulls it down the ridge and the prior only partly resists. The
*sum* recovered accurately over the same fit (1.08 against a true 1.10).

So prefer `bs + sigmabias` whenever you interpret a threshold or compare
one across conditions, and treat the split between the two as
weakly-determined.


## Model Comparison

### Model Fit

```{r}
#| echo: false

path <- "https://raw.github.com/DominiqueMakowski/cogmod/main/vignettes/models/"

m_ddm <- readRDS(url(paste0(path, "m_ddm.rds")))
m_lba <- readRDS(url(paste0(path, "m_lba2.rds")))
m_lnr <- readRDS(url(paste0(path, "m_lnr.rds")))
m_rdm <- readRDS(url(paste0(path, "m_rdm.rds")))
```

```{r}
#| code-fold: false

loo::loo_compare(m_ddm, m_lba, m_lnr, m_rdm) |>
  report::report()
```

Two features of that ranking are worth dwelling on, because neither is
what one might expect.

The first is that the RDM finishes behind the LBA, despite being the more
parsimonious race - it buys its economy by dropping the between-trial
drift variability that the LBA carries in `sigmazero`/`sigmaone`. On this
participant, that variability earns its keep. This is exactly the kind of
question a common interface is for: the claim that within-trial noise
alone suffices is an empirical one, and here it is tested rather than
assumed.

The second is that the DDM comes last, even though it is the only model
here allowed to move its threshold across conditions (`bs ~ Condition`),
which is the mechanism the speed instructions are believed to act on. The
explanation is that its deficit lies on a different axis entirely. We
fitted the *simple* 4-parameter DDM, with `sigmadrift`, `sigmabias` and
`sigmatau` all fixed to `0`, and a DDM stripped of between-trial
variability can only produce errors that are *faster* than correct
responses. In this dataset they are slightly slower (median 0.533 s
against 0.519 s). Sampling from each fitted model makes the mismatch
plain: the DDM predicts errors 0.11 s faster than correct responses (95%
CI `[-0.13, -0.08]`), while the observed difference is `+0.014` s - far
outside that interval. The races each carry a separate drift for the
error accumulator and so are not forced into that prediction: the RDM
puts the gap at `+0.006` `[-0.015, +0.031]`, the LBA at `+0.004`
`[-0.021, +0.036]` and the LNR at `-0.008` `[-0.036, +0.021]`, all three
comfortably covering the observed value. Getting the condition mechanism
right does not compensate
for being unable to represent the error distribution at all, and the
remedy is to free the variability parameters rather than to abandon the
DDM.

### Sampling Duration

As with the RT-only models, we summarize each model's sampling duration by
the median time per chain. Choice+RT models are considerably more
expensive to sample than RT-only models (see the
[RT-only Models](rt_models.html) vignette): the DDM relies on Stan's
`wiener_lpdf`, which is comparatively slow, while the LBA, LNR and RDM
likelihoods involve evaluating both a "winner" density and a "loser"
survival function for every observation. Among the three races, the LNR is
by far the cheapest, since its density and survival are just LogNormal
ones, while the RDM and the LBA each need several normal CDF evaluations
per observation and land in the same, much slower, range.

```{r}
duration <- rbind(
  data_modify(attributes(m_ddm$fit)$metadata$time$chain, Model = "DDM"),
  data_modify(attributes(m_lba$fit)$metadata$time$chain, Model = "LBA"),
  data_modify(attributes(m_lnr$fit)$metadata$time$chain, Model = "LNR"),
  data_modify(attributes(m_rdm$fit)$metadata$time$chain, Model = "RDM")
) |>
  data_modify(Model = factor(Model, levels = c("LNR", "DDM", "RDM", "LBA")))

duration_median <- aggregate(total ~ Model, data = duration, FUN = median)

duration_median |>
  ggplot(aes(x = Model, y = total, fill = Model)) +
  geom_col() +
  geom_text(aes(label = round(total, 1)), vjust = -0.5, size = 3.5) +
  labs(y = "Median Sampling Duration per Chain (s)") +
  scale_fill_material_d(guide = "none") +
  theme_minimal()
```

### Posterior Predictive Check

```{r}
#| code-fold: true

.density_rt_response <- function(rt, response, length.out = 100) {
  rt_error <- rt[response == 0]
  rt_correct <- rt[response == 1]
  xaxis <- seq(0, max(rt_error, rt_correct) * 1.1, length.out = length.out)

  insight::check_if_installed("logspline")
  rbind(
    data.frame(x = xaxis,
               y = logspline::dlogspline(xaxis, logspline::logspline(rt_error)),
               response = 0),
    data.frame(x = xaxis,
               y = -logspline::dlogspline(xaxis, logspline::logspline(rt_correct)),
               response = 1)
  )
}

density_rt_response <- function(data, rt = "RT", response = "Error", by = NULL, length.out = 100) {
  if (is.null(by)) {
    out <- .density_rt_response(data[[rt]], data[[response]], length.out = length.out)
  } else {
    out <- sapply(split(data, data[[by]]), function(x) {
      d <- .density_rt_response(x[[rt]], x[[response]], length.out = length.out)
      d[[by]] <- x[[by]][1]
      d
    }, simplify = FALSE)
    out <- do.call(rbind, out)
    out[[by]] <- as.factor(out[[by]])
  }
  out[[response]] <- as.factor(out[[response]])
  row.names(out) <- NULL
  out
}

pred <- rbind(
  estimate_prediction(m_ddm, data = df, iterations = 50, keep_iterations = TRUE) |>
    reshape_iterations() |>
    data_modify(Model = "DDM"),
  estimate_prediction(m_lba, data = df, iterations = 50, keep_iterations = TRUE) |>
    reshape_iterations() |>
    data_modify(Model = "LBA"),
  estimate_prediction(m_lnr, data = df, iterations = 50, keep_iterations = TRUE) |>
    reshape_iterations() |>
    data_modify(Model = "LNR")
) |>
  datawizard::data_select(select = c("Row", "Component", "iter_value", "iter_group", "iter_index", "Model")) |>
  datawizard::data_to_wide(id_cols = c("Row", "iter_group", "Model"), values_from = "iter_value", names_from = "Component")

# dat <- rbind(
#   data_modify(density_rt_response(pred[pred$Model == "DDM", ], by = "iter_group"), Model = "DDM"),
#   data_modify(density_rt_response(pred[pred$Model == "LBA", ], by = "iter_group"), Model = "LBA"),
#   data_modify(density_rt_response(pred[pred$Model == "LNR", ], by = "iter_group"), Model = "LNR")
# )
# 
# p <- ggplot(df, aes(x = rt)) +
#   geom_histogram(data = df[df$response == 1, ], aes(y = after_stat(density)), fill = "darkgreen", bins = 100) +
#   geom_histogram(data = df[df$response == 0, ], aes(y = after_stat(-density)), fill = "darkred", bins = 100) +
#   geom_line(data = dat, aes(x = x, y = y, group = interaction(response, iter_group)), color = "grey30", alpha = 0.1) +
#   facet_wrap(~Model) +
#   coord_cartesian(xlim = c(0, 2)) +
#   labs(x = "RT (s)", y = "Density (top = Correct, bottom = Error)") +
#   theme_minimal()
# p
```

```{r}
#| echo: false
#| eval: false

# ggsave("../man/figures/decision_making1.png", p, width = 10, height = 5, bg = "white")
```
