---
title: "Assessing Reliability and Interindividual Variability"
vignette: >
  %\VignetteIndexEntry{Assessing Reliability}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
format:
  html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(cogmod)
library(easystats)
library(dplyr)
library(ggplot2)
library(brms)
library(cmdstanr)

options(mc.cores = parallel::detectCores() - 2)
```

Imagine we have designed a brand new cognitive task. It has three conditions -
**A**, **B** and **C** - and our ambition is to use it in a **clinical
context**: to quantify something about *individuals*, so that a patient's score
can be related to their symptoms, compared to a norm, or tracked over time.

The first thing we do is run a validation study and test whether the conditions
affect reaction times. As we will see, only one of them turns out to have a
significant effect - and the tempting conclusion is that this effect is the one
the task actually captures, and therefore the one we should be computing for
each patient.

That conclusion would be premature, and most likely wrong. An average effect
tells us what the manipulation does *to the group*. It says nothing about
whether it does the same thing to everybody, and nothing about whether we can
measure it in a single person with any precision. Yet those two properties -
**interindividual variability** and **reliability** - are exactly what
determines whether a task can serve as an individual-level instrument, and they
have to be assessed on their own terms.

## Data Description

Our validation study is a typical repeated-measures design: 30 participants
each perform 80 trials in each of the three conditions, and we record their
reaction times.




```{r}
#| code-fold: true

# Because we are simulating the data, we get to decide what is true, and we make
# the two effects *deliberately* different:
# 
# - **A vs. B**: a small but extremely **consistent** effect. Every participant
#   is slowed down by about the same amount (the between-participant SD of the
#   effect is tiny, 0.01). There is essentially no interindividual variability.
# - **A vs. C**: an effect with a population mean of **exactly zero**, but a very
#   large between-participant SD (0.30). Some participants are dramatically
#   slowed down by condition C, others are dramatically speeded up.

set.seed(14)

n_participants <- 30
n_trials <- 80  # Per condition

participants <- data.frame(
  Participant = sprintf("S%02d", 1:n_participants),
  # Overall speed: participants differ substantially from one another
  Intercept = rnorm(n_participants, mean = log(0.6), sd = 0.20),
  # A vs. B: small effect, (almost) identical for everybody
  Effect_B = rnorm(n_participants, mean = 0.10, sd = 0.01),
  # A vs. C: no effect on average, but huge interindividual variability
  Effect_C = rnorm(n_participants, mean = 0.00, sd = 0.30)
)

sim <- expand.grid(
  Trial = 1:n_trials,
  Condition = c("A", "B", "C"),
  Participant = participants$Participant,
  stringsAsFactors = FALSE
) |>
  left_join(participants, by = "Participant") |>
  mutate(
    # Trial-level mean on the log scale
    meanlog = Intercept +
      if_else(Condition == "B", Effect_B, 0) +
      if_else(Condition == "C", Effect_C, 0),
    # Generate RTs, adding within-participant (trial-to-trial) noise
    RT = rlnorm(n(), meanlog = meanlog, sdlog = 0.22),
    Participant = factor(Participant),
    Condition = factor(Condition)
  ) |>
  arrange(Participant, Condition, Trial) |>
  select(Participant, Trial, Condition, RT)
```

Looking at the raw distributions, the three conditions are nearly
indistinguishable, and nothing suggests that B and C differ in any interesting
way.

```{r}
ggplot(sim, aes(x = RT, fill = Condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("A" = "grey40", "B" = "purple", "C" = "orange")) +
  labs(x = "Reaction Time (s)", y = "Density") +
  theme_minimal()
```

## Empirical Estimates

Let us now compute the score we would hand to a clinician: for each
participant, average the RTs within each condition and subtract, giving an
"empirical" effect of B - A and of C - A (expressed here in milliseconds).

```{r}
#| code-fold: false

empirical <- sim |>
  summarise(RT = mean(RT), .by = c(Participant, Condition)) |>
  reframe(
    Effect = c("B - A", "C - A"),
    Difference = c(
      RT[Condition == "B"] - RT[Condition == "A"],
      RT[Condition == "C"] - RT[Condition == "A"]
    ) * 1000,
    .by = Participant
  )

empirical |>
  summarise(
    Mean = mean(Difference),
    SD = sd(Difference),
    Min = min(Difference),
    Max = max(Difference),
    .by = Effect
  )
```

The two effects could not be more different, and yet **the number we usually
report - the mean - is blind to it**. The B - A effect is about +74 ms on
average, with a between-participant SD of ~31 ms. The C - A effect has a mean
of essentially zero (+0.4 ms) - but a between-participant SD of ~171 ms, and it
ranges from -240 ms to +408 ms depending on which participant we look at.

```{r}
# Sort participants by the size of their C - A effect
order_C <- empirical |>
  filter(Effect == "C - A") |>
  arrange(Difference) |>
  pull(Participant)

empirical |>
  mutate(Participant = factor(Participant, levels = order_C)) |>
  ggplot(aes(x = Difference, y = Participant, color = Effect)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 2.5) +
  facet_wrap(~Effect) +
  scale_color_manual(values = c("B - A" = "purple", "C - A" = "orange")) +
  labs(x = "RT difference (ms)", y = "Participant") +
  theme_minimal() +
  theme(legend.position = "none")
```

The picture is unambiguous (note that participants are ordered by the size of
their C - A effect, so that a given row refers to the same person in both
panels). On the left, every single participant shows the same small slowing in
condition B: the effect is *reliable* in the sense of being reproducible across
people, but there is almost nothing to distinguish one participant from
another. On the right, participants are spread all over the place, with some
being massively slower and others massively faster in condition C - and these
differences cancel out to nothing at the group level.

This is bad news for our original plan. Condition B - the one that reached
significance, and on which we were about to build our clinical index - gives
essentially the same score to everybody, and a measure that does not vary
between people cannot possibly correlate with symptoms, separate patients from
controls, or track change. Condition C - the "null" effect that we would have
discarded - is the one carrying the individual differences we are after.

> **Careful: empirical estimates are noisy!**

These per-participant differences are *not* clean measurements of each
participant's true effect. Each one is computed from a finite number of trials
and therefore carries measurement error. Note for instance that the B - A
effects show an apparent spread of ~31 ms even though we generated them with a
between-participant SD of only ~6 ms (0.01 on the log scale): almost all of
that visible spread is noise, not real interindividual variability.
Disentangling the two is precisely what the model below does.


<!-- ## Modeling -->

<!-- We now fit a **mixed LogNormal model**, which mirrors the way the data were -->
<!-- generated: the effect of `Condition` is estimated at the population level -->
<!-- (fixed effects) *and* allowed to vary across participants (random slopes). -->

<!-- ```{r} -->
<!-- #| eval: false -->
<!-- #| code-fold: false -->

<!-- f <- bf(RT ~ Condition + (Condition | Participant)) -->

<!-- m <- brm(f, -->
<!--   data = sim, -->
<!--   family = lognormal(), -->
<!--   init = 0, -->
<!--   chains = 4, iter = 1000, backend = "cmdstanr" -->
<!-- ) -->

<!-- saveRDS(m, file = "models/m_reliability.rds") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| echo: false -->

<!-- path <- "https://raw.github.com/DominiqueMakowski/cogmod/main/vignettes/models/" -->

<!-- m <- readRDS(url(paste0(path, "m_reliability.rds"))) -->
<!-- ``` -->

<!-- The random-effects part `(Condition | Participant)` is the key ingredient: it -->
<!-- tells the model that each participant has their own intercept *and* their own -->
<!-- condition effects, and it estimates how much these vary. Note also that we do -->
<!-- not need to transform the RTs: `family = lognormal()` handles the skew -->
<!-- directly, and the coefficients are simply expressed on the log scale. -->

<!-- Let us first look at the **fixed effects**, i.e., the population-level answer -->
<!-- to "is there an effect of condition?". -->

<!-- ```{r} -->
<!-- #| code-fold: false -->

<!-- parameters(m, effects = "fixed") -->
<!-- ``` -->

<!-- This reproduces the classic conclusion: a clear effect of condition B -->
<!-- (0.10 on the log scale, 95% CI [0.09, 0.12], *pd* = 100%), and nothing at all -->
<!-- for condition C (-0.02, 95% CI [-0.12, 0.08], *pd* = 65%), whose credible -->
<!-- interval is comfortably centered on zero. -->

<!-- The crucial addition is the **random effects**, which quantify how much -->
<!-- participants differ from each other on each of these parameters: -->

<!-- ```{r} -->
<!-- #| code-fold: false -->

<!-- parameters(m, effects = "random_variance") -->
<!-- ``` -->

<!-- Here the two effects finally become distinguishable. The SD of the `ConditionB` -->
<!-- slope is 0.01, i.e., essentially nothing: participants barely differ in how -->
<!-- much B affects them. The SD of the `ConditionC` slope, in contrast, is 0.26 - -->
<!-- more than twenty times larger, and very close to the value of 0.30 we -->
<!-- simulated. This is the model formally telling us that *there is a lot going on -->
<!-- in condition C, it just does not go in the same direction for everybody*. -->
<!-- Note also that the model recovers the true between-participant SD of the B -->
<!-- effect (~0.01) rather than the inflated ~31 ms spread of the raw empirical -->
<!-- differences: the measurement noise has been correctly assigned to the residual -->
<!-- term instead of being mistaken for interindividual variability. -->

<!-- ## Group-Level Indices Extraction -->

<!-- Beyond knowing that participants differ, we often want to know *how* each -->
<!-- individual differs - for instance to correlate these individual effects with -->
<!-- another variable (a questionnaire score, a diagnosis, another task). These -->
<!-- participant-level deviations can be extracted with -->
<!-- `modelbased::estimate_grouplevel()`. -->

<!-- ```{r} -->
<!-- #| code-fold: false -->

<!-- random <- estimate_grouplevel(m) -->

<!-- head(random) -->
<!-- ``` -->

<!-- Each row is one participant's deviation from the population-level (fixed) -->
<!-- effect, together with its uncertainty. We can visualize all of them at once: -->

<!-- ```{r} -->
<!-- #| fig-height: 8 -->

<!-- plot(random) + -->
<!--   theme_minimal() -->
<!-- ``` -->

<!-- The contrast between the two effects is now expressed as model parameters -->
<!-- rather than as raw averages (note that each panel has its own x-axis scale). -->
<!-- For `ConditionB`, all the individual deviations are packed within ±0.03 of -->
<!-- zero and every single credible interval includes zero: the model concludes -->
<!-- that, once measurement noise is accounted for, participants essentially share -->
<!-- the same effect. For `ConditionC`, the deviations span -0.45 to +0.56 and most -->
<!-- of them clearly exclude zero - these are genuine, statistically supported -->
<!-- individual differences. -->

<!-- Because these estimates come from a model, they are also **shrunk** towards -->
<!-- the population mean in proportion to how noisy each participant's data is: -->
<!-- participants with fewer or more variable trials are pulled more strongly -->
<!-- towards the average. This is a feature, not a bug - it prevents us from -->
<!-- over-interpreting extreme values that the raw empirical differences produce by -->
<!-- chance, and it is why model-based individual scores are generally more -->
<!-- reliable than difference scores computed by hand. -->

<!-- How much this matters depends on how much data each participant provides. Here, -->
<!-- with 80 trials per condition, each individual effect is already well estimated, -->
<!-- so the model-based indices and the raw empirical differences agree almost -->
<!-- perfectly (*r* = .98) and the plot below is essentially a straight line. The -->
<!-- two would come apart - with the model-based version being the more trustworthy -->
<!-- of the two - in the more realistic situation of fewer trials, noisier -->
<!-- participants, or unbalanced designs. -->

<!-- ```{r} -->
<!-- #| code-fold: true -->

<!-- # Compare the model-based individual effects to the raw empirical ones -->
<!-- comparison <- merge( -->
<!--   data_filter(as.data.frame(random), Parameter == "ConditionC") |> -->
<!--     data_select(c("Level", "Median")) |> -->
<!--     data_rename("Level", "Participant"), -->
<!--   data_filter(empirical, Effect == "C - A") |> -->
<!--     data_select(c("Participant", "Difference")) -->
<!-- ) -->

<!-- ggplot(comparison, aes(x = Difference, y = Median)) + -->
<!--   geom_hline(yintercept = 0, linetype = "dashed") + -->
<!--   geom_vline(xintercept = 0, linetype = "dashed") + -->
<!--   geom_point(size = 2.5, color = "darkred") + -->
<!--   labs( -->
<!--     x = "Empirical effect: mean(C) - mean(A), in ms", -->
<!--     y = "Model-based individual effect (log scale)" -->
<!--   ) + -->
<!--   theme_minimal() -->
<!-- ``` -->

<!-- ## Conclusion -->

<!-- The take-home message is that **an average effect and an interindividual -->
<!-- difference are two different things**, and that a study designed to detect the -->
<!-- first can be blind to the second. A significant fixed effect does not imply -->
<!-- that the task is useful to characterize individuals, and a null fixed effect -->
<!-- does not imply that nothing is happening. -->

<!-- Mixed models give access to both at once: the fixed effects answer the -->
<!-- traditional group-level question, the random-effect SDs tell you whether there -->
<!-- is any interindividual variability to speak of, and the group-level indices -->
<!-- give you per-participant estimates that are usable as individual scores. -->
