---
title: "Assessing Reliability and Interindividual Variability"
vignette: >
  %\VignetteIndexEntry{Assessing Reliability}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
format:
  html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(cogmod)
library(easystats)
library(ggplot2)
library(brms)
library(cmdstanr)

options(mc.cores = parallel::detectCores() - 2)
```

Experimental psychology has traditionally been interested in **average
effects**: does this manipulation change reaction times *in general*? But an
average effect is only one property of an experimental task, and arguably not
always the most interesting one. A manipulation can produce a rock-solid
average effect that is identical in everybody - in which case there is nothing
to correlate with anything, and the task is useless as an individual
differences measure. Conversely, a manipulation can have *no* average effect
whatsoever while producing large and systematic differences *between*
individuals - which makes it a potentially excellent marker of something, even
though the classic group-level test would declare it a failure.

This vignette illustrates that distinction on simulated data, and shows how to
recover participant-level effects from a mixed model.

## Data Description

We simulate a typical repeated-measures experiment: 30 participants each
perform 80 trials in **three conditions** (A, B and C), and we record their
reaction times.

The data are generated from a **LogNormal** distribution, which - as discussed
in the [RT models vignette](https://dominiquemakowski.github.io/cogmod/articles/rt_models.html) -
is a much more sensible generative model for RTs than a Gaussian. Everything
happens on the log scale: each participant has their own baseline speed
(`Intercept`), and their own effect of condition B and C relative to A.

The two effects are built to be *deliberately* different:

- **A vs. B**: a small but extremely **consistent** effect. Every participant
  is slowed down by about the same amount (the between-participant SD of the
  effect is tiny, 0.01). There is essentially no interindividual variability.
- **A vs. C**: an effect with a population mean of **exactly zero**, but a very
  large between-participant SD (0.30). Some participants are dramatically
  slowed down by condition C, others are dramatically speeded up.

```{r}
#| code-fold: true

set.seed(14)

n_participants <- 30
n_trials <- 80  # Per condition

participants <- data.frame(
  Participant = sprintf("S%02d", 1:n_participants),
  # Overall speed: participants differ substantially from one another
  Intercept = rnorm(n_participants, mean = log(0.6), sd = 0.20),
  # A vs. B: small effect, (almost) identical for everybody
  Effect_B = rnorm(n_participants, mean = 0.10, sd = 0.01),
  # A vs. C: no effect on average, but huge interindividual variability
  Effect_C = rnorm(n_participants, mean = 0.00, sd = 0.30)
)

sim <- expand.grid(
  Trial = 1:n_trials,
  Condition = c("A", "B", "C"),
  Participant = participants$Participant,
  stringsAsFactors = FALSE
)
sim <- merge(sim, participants, by = "Participant")

# Trial-level mean on the log scale
sim$meanlog <- sim$Intercept +
  ifelse(sim$Condition == "B", sim$Effect_B, 0) +
  ifelse(sim$Condition == "C", sim$Effect_C, 0)

# Generate RTs, adding within-participant (trial-to-trial) noise
sim$RT <- rlnorm(nrow(sim), meanlog = sim$meanlog, sdlog = 0.22)

sim <- sim[order(sim$Participant, sim$Condition, sim$Trial), ]
sim <- data_select(sim, c("Participant", "Trial", "Condition", "RT"))
sim$Participant <- factor(sim$Participant)
sim$Condition <- factor(sim$Condition)

head(sim)
```

Looking at the raw distributions, the three conditions are nearly
indistinguishable, and nothing suggests that B and C differ in any interesting
way.

```{r}
ggplot(sim, aes(x = RT, fill = Condition)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("A" = "grey40", "B" = "darkgreen", "C" = "darkred")) +
  labs(x = "Reaction Time (s)", y = "Density") +
  theme_minimal()
```

<!-- ## Empirical Estimates -->

<!-- Before modelling anything, we can compute the most straightforward possible -->
<!-- index of each participant's effect: average the RTs within each condition, and -->
<!-- subtract. This gives us, for every participant, an "empirical" effect of -->
<!-- B - A and of C - A (expressed here in milliseconds). -->

<!-- ```{r} -->
<!-- #| code-fold: false -->

<!-- # Average RT per participant and condition -->
<!-- persub <- aggregate(RT ~ Participant + Condition, data = sim, FUN = mean) -->
<!-- persub <- data_to_wide(persub, id_cols = "Participant", -->
<!--                        names_from = "Condition", values_from = "RT") -->

<!-- empirical <- data.frame( -->
<!--   Participant = rep(persub$Participant, 2), -->
<!--   Effect = rep(c("B - A", "C - A"), each = nrow(persub)), -->
<!--   Difference = c(persub$B - persub$A, persub$C - persub$A) * 1000 -->
<!-- ) -->

<!-- # Summary of the two effects across participants -->
<!-- aggregate(Difference ~ Effect, data = empirical, FUN = function(x) { -->
<!--   c(Mean = mean(x), SD = sd(x), Min = min(x), Max = max(x)) -->
<!-- }) -->
<!-- ``` -->

<!-- The two effects could not be more different, and yet **only one of the two -->
<!-- numbers usually reported - the mean - captures it**. The B - A effect is about -->
<!-- +74 ms on average, with a between-participant SD of ~31 ms. The C - A effect -->
<!-- has a mean of essentially zero (+0.4 ms) - but a between-participant SD of -->
<!-- ~171 ms, and it ranges from -240 ms to +408 ms depending on who you look at. -->

<!-- ```{r} -->
<!-- ggplot(empirical, aes(x = reorder(Participant, Difference), y = Difference, color = Effect)) + -->
<!--   geom_hline(yintercept = 0, linetype = "dashed") + -->
<!--   geom_point(size = 2.5) + -->
<!--   facet_wrap(~Effect, scales = "free_x") + -->
<!--   scale_color_manual(values = c("B - A" = "darkgreen", "C - A" = "darkred")) + -->
<!--   labs(x = "Participant (sorted by effect)", y = "RT difference (ms)") + -->
<!--   theme_minimal() + -->
<!--   theme(axis.text.x = element_blank(), legend.position = "none") -->
<!-- ``` -->

<!-- The picture is unambiguous. On the left, every single participant shows the -->
<!-- same small slowing in condition B: the effect is *reliable* in the sense of -->
<!-- being reproducible across people, but there is almost nothing to distinguish -->
<!-- one participant from another. On the right, participants are spread all over -->
<!-- the place, with some being massively slower and others massively faster in -->
<!-- condition C - and these differences cancel out to nothing at the group level. -->

<!-- A traditional analysis would report a significant effect of B and a null -->
<!-- effect of C, and would stop there. Yet if you were looking for a task that can -->
<!-- be used to measure something *about individuals* (a trait, a clinical marker, -->
<!-- a correlate of some other variable), the "null" effect C is the interesting -->
<!-- one, and the "significant" effect B is the useless one. -->

<!-- ::: {.callout-note title="Careful: empirical estimates are noisy"} -->

<!-- These per-participant differences are *not* clean measurements of each -->
<!-- participant's true effect. Each one is computed from a finite number of trials -->
<!-- and therefore carries measurement error. Note for instance that the B - A -->
<!-- effects show an apparent spread of ~31 ms even though we generated them with a -->
<!-- between-participant SD of only ~6 ms (0.01 on the log scale): almost all of -->
<!-- that visible spread is noise, not real interindividual variability. -->
<!-- Disentangling the two is precisely what the model below does. -->

<!-- ::: -->

<!-- ## Modeling -->

<!-- We now fit a **mixed LogNormal model**, which mirrors the way the data were -->
<!-- generated: the effect of `Condition` is estimated at the population level -->
<!-- (fixed effects) *and* allowed to vary across participants (random slopes). -->

<!-- ```{r} -->
<!-- #| eval: false -->
<!-- #| code-fold: false -->

<!-- f <- bf(RT ~ Condition + (Condition | Participant)) -->

<!-- m <- brm(f, -->
<!--   data = sim, -->
<!--   family = lognormal(), -->
<!--   init = 0, -->
<!--   chains = 4, iter = 1000, backend = "cmdstanr" -->
<!-- ) -->

<!-- saveRDS(m, file = "models/m_reliability.rds") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| echo: false -->

<!-- path <- "https://raw.github.com/DominiqueMakowski/cogmod/main/vignettes/models/" -->

<!-- m <- readRDS(url(paste0(path, "m_reliability.rds"))) -->
<!-- ``` -->

<!-- The random-effects part `(Condition | Participant)` is the key ingredient: it -->
<!-- tells the model that each participant has their own intercept *and* their own -->
<!-- condition effects, and it estimates how much these vary. Note also that we do -->
<!-- not need to transform the RTs: `family = lognormal()` handles the skew -->
<!-- directly, and the coefficients are simply expressed on the log scale. -->

<!-- Let us first look at the **fixed effects**, i.e., the population-level answer -->
<!-- to "is there an effect of condition?". -->

<!-- ```{r} -->
<!-- #| code-fold: false -->

<!-- parameters(m, effects = "fixed") -->
<!-- ``` -->

<!-- This reproduces the classic conclusion: a clear effect of condition B -->
<!-- (0.10 on the log scale, 95% CI [0.09, 0.12], *pd* = 100%), and nothing at all -->
<!-- for condition C (-0.02, 95% CI [-0.12, 0.08], *pd* = 65%), whose credible -->
<!-- interval is comfortably centered on zero. -->

<!-- The crucial addition is the **random effects**, which quantify how much -->
<!-- participants differ from each other on each of these parameters: -->

<!-- ```{r} -->
<!-- #| code-fold: false -->

<!-- parameters(m, effects = "random_variance") -->
<!-- ``` -->

<!-- Here the two effects finally become distinguishable. The SD of the `ConditionB` -->
<!-- slope is 0.01, i.e., essentially nothing: participants barely differ in how -->
<!-- much B affects them. The SD of the `ConditionC` slope, in contrast, is 0.26 - -->
<!-- more than twenty times larger, and very close to the value of 0.30 we -->
<!-- simulated. This is the model formally telling us that *there is a lot going on -->
<!-- in condition C, it just does not go in the same direction for everybody*. -->
<!-- Note also that the model recovers the true between-participant SD of the B -->
<!-- effect (~0.01) rather than the inflated ~31 ms spread of the raw empirical -->
<!-- differences: the measurement noise has been correctly assigned to the residual -->
<!-- term instead of being mistaken for interindividual variability. -->

<!-- ## Group-Level Indices Extraction -->

<!-- Beyond knowing that participants differ, we often want to know *how* each -->
<!-- individual differs - for instance to correlate these individual effects with -->
<!-- another variable (a questionnaire score, a diagnosis, another task). These -->
<!-- participant-level deviations can be extracted with -->
<!-- `modelbased::estimate_grouplevel()`. -->

<!-- ```{r} -->
<!-- #| code-fold: false -->

<!-- random <- estimate_grouplevel(m) -->

<!-- head(random) -->
<!-- ``` -->

<!-- Each row is one participant's deviation from the population-level (fixed) -->
<!-- effect, together with its uncertainty. We can visualize all of them at once: -->

<!-- ```{r} -->
<!-- #| fig-height: 8 -->

<!-- plot(random) + -->
<!--   theme_minimal() -->
<!-- ``` -->

<!-- The contrast between the two effects is now expressed as model parameters -->
<!-- rather than as raw averages (note that each panel has its own x-axis scale). -->
<!-- For `ConditionB`, all the individual deviations are packed within ±0.03 of -->
<!-- zero and every single credible interval includes zero: the model concludes -->
<!-- that, once measurement noise is accounted for, participants essentially share -->
<!-- the same effect. For `ConditionC`, the deviations span -0.45 to +0.56 and most -->
<!-- of them clearly exclude zero - these are genuine, statistically supported -->
<!-- individual differences. -->

<!-- Because these estimates come from a model, they are also **shrunk** towards -->
<!-- the population mean in proportion to how noisy each participant's data is: -->
<!-- participants with fewer or more variable trials are pulled more strongly -->
<!-- towards the average. This is a feature, not a bug - it prevents us from -->
<!-- over-interpreting extreme values that the raw empirical differences produce by -->
<!-- chance, and it is why model-based individual scores are generally more -->
<!-- reliable than difference scores computed by hand. -->

<!-- How much this matters depends on how much data each participant provides. Here, -->
<!-- with 80 trials per condition, each individual effect is already well estimated, -->
<!-- so the model-based indices and the raw empirical differences agree almost -->
<!-- perfectly (*r* = .98) and the plot below is essentially a straight line. The -->
<!-- two would come apart - with the model-based version being the more trustworthy -->
<!-- of the two - in the more realistic situation of fewer trials, noisier -->
<!-- participants, or unbalanced designs. -->

<!-- ```{r} -->
<!-- #| code-fold: true -->

<!-- # Compare the model-based individual effects to the raw empirical ones -->
<!-- comparison <- merge( -->
<!--   data_filter(as.data.frame(random), Parameter == "ConditionC") |> -->
<!--     data_select(c("Level", "Median")) |> -->
<!--     data_rename("Level", "Participant"), -->
<!--   data_filter(empirical, Effect == "C - A") |> -->
<!--     data_select(c("Participant", "Difference")) -->
<!-- ) -->

<!-- ggplot(comparison, aes(x = Difference, y = Median)) + -->
<!--   geom_hline(yintercept = 0, linetype = "dashed") + -->
<!--   geom_vline(xintercept = 0, linetype = "dashed") + -->
<!--   geom_point(size = 2.5, color = "darkred") + -->
<!--   labs( -->
<!--     x = "Empirical effect: mean(C) - mean(A), in ms", -->
<!--     y = "Model-based individual effect (log scale)" -->
<!--   ) + -->
<!--   theme_minimal() -->
<!-- ``` -->

<!-- ## Conclusion -->

<!-- The take-home message is that **an average effect and an interindividual -->
<!-- difference are two different things**, and that a study designed to detect the -->
<!-- first can be blind to the second. A significant fixed effect does not imply -->
<!-- that the task is useful to characterize individuals, and a null fixed effect -->
<!-- does not imply that nothing is happening. -->

<!-- Mixed models give access to both at once: the fixed effects answer the -->
<!-- traditional group-level question, the random-effect SDs tell you whether there -->
<!-- is any interindividual variability to speak of, and the group-level indices -->
<!-- give you per-participant estimates that are usable as individual scores. -->
