---
title: "RT-only Models"
vignette: >
  %\VignetteIndexEntry{RT-only Models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
format: 
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(cogmod)
library(easystats)
library(ggplot2)
library(brms)
library(cmdstanr)

options(mc.cores = parallel::detectCores() - 2)
```


## The Data

For this chapter, we will be using the data from [Wagenmakers et al., (2008)](https://doi.org/10.1016/j.jml.2007.04.006) - Experiment 1 also reanalyzed by [Heathcote & Love (2012)](https://doi.org/10.3389/fpsyg.2012.00292), that contains responses and response times for several participants in two conditions (where instructions emphasized either **speed** or **accuracy**). This dataset is bundled directly with `cogmod` as `wagenmakers2008`.
Using the same procedure as the authors, we excluded all trials with uninterpretable response time, i.e., responses that are too fast (<200 ms instead of <180 ms) or too slow (>2 sec instead of >3 sec).

```{r}
#| code-fold: false

set.seed(123)  # For reproducibility

df <- cogmod::wagenmakers2008
df <- df[df$RT > 0.2 & df$Participant %in% c(1, 2, 3), ]

# Show 10 first rows
head(df, 10)
```

<!-- In the previous chapter, we modeled the error rate (the probability of making an error) using a logistic model, and observed that it was higher in the `"Speed"` condition.  -->
<!-- But how about speed?  -->
We are going to first take interest in the response times (RT) of **Correct** answers only (as we can assume that errors are underpinned by a different *generative process*). 

```{r}
#| output: false

df <- df[df$Error == 0, ]
```


```{r}
ggplot(df, aes(x = RT, fill = Condition)) +
  geom_histogram(bins = 120, alpha = 0.8, position = "identity") +
  scale_fill_manual(values = c("darkgreen", "darkred")) +
  theme_minimal()
```

## Models

### Normal

A basic linear model.

```{r}
#| eval: false
#| code-fold: true

f <- bf(
  RT ~ Condition
)

m_normal <- brm(f,
  data = df, 
  init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_normal <- brms::add_criterion(m_normal, "loo") 

saveRDS(m_normal, file = "models/m_normal.rds")
```



### ExGaussian

The Ex-Gaussian distribution is a popular choice for RT data, as it separates
the central portion of the distribution - captured by the Gaussian parameters
`mu` and `sigma` - from the positively skewed tail, captured by the
exponential parameter `tau`. However, `brms`'s native `exgaussian()` family
does **not** use this "classical" parameterization familiar to experimental
psychologists: its `mu` indexes the mean of the *entire* distribution (i.e.,
Gaussian + exponential combined) rather than the location of the Gaussian
component alone. This matters because a change in the Gaussian location and
an opposite change in the exponential tail can cancel out at the level of the
overall mean, so effects estimated on `brms`'s default `mu` can lead to
different (and potentially incorrect) inferences about the underlying
process than effects estimated on the classical `mu`.

For this reason, `cogmod` provides its own `rt_exgaussian()` custom family
(internally relying on Stan's `exp_mod_normal` distribution), in which `mu`
and `sigma` are the mean and SD of the Gaussian component and `tau` is the
mean of the exponential tail - directly matching the classical
parameterization.

```{r}
#| eval: false
#| code-fold: true

f <- bf(
  RT ~ Condition,
  sigma ~ Condition,
  tau ~ Condition,
  family = rt_exgaussian()
)

m_exgauss <- brm(f,
  data = df, 
  stanvars = rt_exgaussian_stanvars(),
  init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_exgauss <- brms::add_criterion(m_exgauss, "loo") 

saveRDS(m_exgauss, file = "models/m_exgauss.rds")
```


### Shifted LogNormal

```{r}
#| eval: false
#| code-fold: true

f <- bf(
  RT ~ Condition,
  sigma ~ Condition,
  tau ~ Condition,
  minrt = min(df$RT),
  family = rt_lognormal()
)

priors <- brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau") |>
  brms::validate_prior(f, data = df)

m_lognormal <- brm(
  f,
  prior = priors,
  data = df, 
  stanvars = rt_lognormal_stanvars(),
  init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_lognormal <- brms::add_criterion(m_lognormal, "loo") 

saveRDS(m_lognormal, file = "models/m_lognormal.rds")
```

### Inverse Gaussian (Shifted Wald)

```{r}
#| eval: false
#| code-fold: true

f <- bf(
  RT ~ Condition,
  bs ~ Condition,
  tau ~ Condition,
  minrt = min(df$RT),
  family = rt_invgaussian()
)

priors <- brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau") |>
  brms::validate_prior(f, data = df)

m_wald <- brm(
  f,
  prior = priors,
  data = df, 
  stanvars = rt_invgaussian_stanvars(),
  init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_wald <- brms::add_criterion(m_wald, "loo") 

saveRDS(m_wald, file = "models/m_wald.rds")
```


### Weibull

```{r}
#| eval: false
#| code-fold: true

f <- bf(
  RT ~ Condition,
  sigma ~ Condition,
  tau ~ Condition,
  minrt = min(df$RT),
  family = rt_weibull()
)

priors <- brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau") |>
  brms::validate_prior(f, data = df)

m_weibull <- brm(
  f,
  prior = priors,
  data = df, 
  stanvars = rt_weibull_stanvars(),
  init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_weibull <- brms::add_criterion(m_weibull, "loo") 

saveRDS(m_weibull, file = "models/m_weibull.rds")
```


### LogWeibull (Shifted Gumbel)

```{r}
#| eval: false
#| code-fold: true

f <- bf(
  RT ~ Condition,
  sigma ~ Condition,
  tau ~ Condition,
  minrt = min(df$RT),
  family = rt_logweibull()
)

priors <- brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau") |>
  brms::validate_prior(f, data = df)

m_logweibull <- brm(
  f,
  prior = priors,
  data = df, 
  stanvars = rt_logweibull_stanvars(),
  init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_logweibull <- brms::add_criterion(m_logweibull, "loo") 

saveRDS(m_logweibull, file = "models/m_logweibull.rds")
```


### Inverse Weibull (Shifted Fréchet)

```{r}
#| eval: false
#| code-fold: true

f <- bf(
  RT ~ Condition,
  sigma ~ Condition,
  tau ~ Condition,
  minrt = min(df$RT),
  family = rt_invweibull()
)

priors <- brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau") |>
  brms::validate_prior(f, data = df)

m_invweibull <- brm(
  f,
  prior = priors,
  data = df, 
  stanvars = rt_invweibull_stanvars(),
  init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_invweibull <- brms::add_criterion(m_invweibull, "loo") 

saveRDS(m_invweibull, file = "models/m_invweibull.rds")
```




### Gamma

```{r}
#| eval: false
#| code-fold: true

f <- bf(
  RT ~ Condition,
  sigma ~ Condition,
  tau ~ Condition,
  minrt = min(df$RT),
  family = rt_gamma()
)

priors <- brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau") |>
  brms::validate_prior(f, data = df)

m_gamma <- brm(
  f,
  prior = priors,
  data = df, 
  stanvars = rt_gamma_stanvars(),
  init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_gamma <- brms::add_criterion(m_gamma, "loo") 

saveRDS(m_gamma, file = "models/m_gamma.rds")
```

### Inverse Gamma

```{r}
#| eval: false
#| code-fold: true

f <- bf(
  RT ~ Condition,
  sigma ~ Condition,
  tau ~ Condition,
  minrt = min(df$RT),
  family = rt_invgamma()
)

priors <- brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau") |>
  brms::validate_prior(f, data = df)

m_invgamma <- brm(
  f,
  prior = priors,
  data = df, 
  stanvars = rt_invgamma_stanvars(),
  init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_invgamma <- brms::add_criterion(m_invgamma, "loo") 

saveRDS(m_invgamma, file = "models/m_invgamma.rds")
```

### Linear Ballistic Accumulator

```{r}
#| eval: false
#| code-fold: true

f <- bf(
  RT ~ Condition,
  sigma = 1,
  sigmabias ~ Condition,
  bs ~ Condition,
  tau ~ Condition,
  minrt = min(df$RT),
  family = rt_lba()
)

priors <- c(
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "mu"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "sigmabias"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "bs"),
  brms::set_prior("normal(0, 1)", class = "Intercept", dpar = "tau")
  )|>
  brms::validate_prior(f, data = df)

m_lba <- brm(
  f,
  prior = priors,
  data = df, 
  stanvars = rt_lba_stanvars(),
  init = 0.5,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_lba <- brms::add_criterion(m_lba, "loo") 

saveRDS(m_lba, file = "models/m_lba.rds")
```




## Model Comparison

### Model Fit

We can compare these models together using the `loo` package, which shows
that CHOCO provides a significantly better fit than the other models.

```{r}
#| echo: false

path <- "https://raw.github.com/DominiqueMakowski/cogmod/main/vignettes/models/"

m_normal <- readRDS(url(paste0(path, "m_normal.rds")))
m_exgauss <- readRDS(url(paste0(path, "m_exgauss.rds")))
m_lognormal <- readRDS(url(paste0(path, "m_lognormal.rds")))
m_wald <- readRDS(url(paste0(path, "m_wald.rds")))
m_weibull <- readRDS(url(paste0(path, "m_weibull.rds")))
m_logweibull <- readRDS(url(paste0(path, "m_logweibull.rds")))
m_invweibull <- readRDS(url(paste0(path, "m_invweibull.rds")))
m_gamma <- readRDS(url(paste0(path, "m_gamma.rds")))
m_invgamma <- readRDS(url(paste0(path, "m_invgamma.rds")))
m_lba <- readRDS(url(paste0(path, "m_lba.rds")))
```


```{r}
#| code-fold: false

loo::loo_compare(m_normal, m_exgauss, m_lognormal, m_wald,
                 m_weibull, m_logweibull, m_invweibull,
                 m_gamma, m_invgamma, m_lba) |>
  parameters(include_ENP = TRUE)
```


### Sampling Duration

Because each model was fit with only 4 chains, a boxplot of the per-chain
sampling times is not very informative. Instead, we summarize each model's
sampling duration by the *median* time per chain, annotated with its value.

As expected, the **Gaussian** model is by far the fastest to sample, since it
relies on `brms`'s built-in (and heavily optimized) Normal likelihood with no
custom Stan code or non-decision time shift involved. At the other end,
the **LBA** is by far the slowest, reflecting the added cost of its
multi-accumulator likelihood. The remaining RT-only models (ExGaussian,
LogNormal, Wald, Weibull, LogWeibull, InvWeibull, Gamma, and InvGamma) are all
relatively comparable to one another, as they share a similar structure
(a simple closed-form density combined with a non-decision time shift).

```{r}
duration <- rbind(
  data_modify(attributes(m_normal$fit)$metadata$time$chain, Model="Gaussian"),
  data_modify(attributes(m_exgauss$fit)$metadata$time$chain, Model="ExGaussian"),
  data_modify(attributes(m_lognormal$fit)$metadata$time$chain, Model="LogNormal"),
  data_modify(attributes(m_wald$fit)$metadata$time$chain, Model="Wald"),
  data_modify(attributes(m_weibull$fit)$metadata$time$chain, Model="Weibull"),
  data_modify(attributes(m_logweibull$fit)$metadata$time$chain, Model="LogWeibull"),
  data_modify(attributes(m_invweibull$fit)$metadata$time$chain, Model="InvWeibull"),
  data_modify(attributes(m_gamma$fit)$metadata$time$chain, Model="Gamma"),
  data_modify(attributes(m_invgamma$fit)$metadata$time$chain, Model="InvGamma"),
  data_modify(attributes(m_lba$fit)$metadata$time$chain, Model="LBA")
) |> 
  data_modify(Model = factor(Model, levels = c("Gaussian", "ExGaussian", "LogNormal", "Wald", "Weibull", "LogWeibull", "InvWeibull", "Gamma", "InvGamma", "LBA")))

duration_median <- aggregate(total ~ Model, data = duration, FUN = median)

duration_median |> 
  ggplot(aes(x = Model, y = total, fill = total)) +
  geom_col() +
  geom_text(aes(label = round(total, 1)), vjust = -0.5, size = 3.5) +
  labs(y = "Median Sampling Duration per Chain (s)", fill = "Duration") +
  scale_fill_gradientn(colors = c("turquoise", "green", "yellow", "gold", "orange", "red", "darkred")) +
  scale_y_sqrt() +
  theme_minimal() 
```


### Posterior Predictive Check

`iterations` controls the actual number of iterations used (e.g., for the point-estimate) and `keep_iterations` the number included.

```{r}
#| code-fold: true


pred <- rbind(
  # estimate_prediction(m_normal, keep_iterations = 50, iterations = 50) |>
  #   reshape_iterations() |>
  #   data_modify(Model = "Normal"),
  estimate_prediction(m_exgauss, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "ExGaussian"),
  estimate_prediction(m_lognormal, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "LogNormal"),
  estimate_prediction(m_wald, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "InvGaussian"),
  estimate_prediction(m_weibull, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "Weibull"),
  estimate_prediction(m_logweibull, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "LogWeibull"),
  estimate_prediction(m_invweibull, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "InvWeibull"),
  estimate_prediction(m_gamma, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "Gamma"),
  estimate_prediction(m_invgamma, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "InvGamma"),
  estimate_prediction(m_lba, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "LBA") |> 
    data_filter(iter_value < 2)
)

p <- pred |> 
  ggplot(aes(x=iter_value)) +
  geom_histogram(data = df, aes(x=RT, y = after_stat(density), fill = Condition), 
                 position = "identity", bins=120, alpha = 0.8) +
  geom_line(aes(color=Model, group=interaction(Condition, iter_group)), stat="density", alpha=0.2) +
  theme_minimal() +
  theme(axis.text.y = element_blank()) +
  facet_wrap(~Model) +
  coord_cartesian(xlim = c(0, 2)) +
  scale_fill_manual(values = c("darkgreen", "darkred")) +
  scale_color_material_d(guide = "none") +
  labs(x = "RT (s)", y = "Distribution")
p
```


```{r}
#| echo: false
#| eval: false

ggsave("../man/figures/rt_models1.png", p, width = 10, height = 8, bg = "white")
```
