---
title: "Subjective Ratings"
vignette: >
  %\VignetteIndexEntry{Subjective Ratings}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
format: 
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(cogmod)
library(easystats)
library(ggplot2)
library(brms)
library(cmdstanr)

options(mc.cores = parallel::detectCores() - 2)
set.seed(33)
```

## Analog Scales

### Simulate Data

```{r}
#| fig-width: 10
#| fig-height: 7
#| warning: false
#| message: false

df <- data.frame()
for(x in seq(0.1, 0.9, by = 0.1)) {
  score <- rchoco(n = 100, p = 0.4 + x / 2, confright = 0.4 + x / 3, 
                  confleft = 1-x, pex = 0.03, bex = 0.6, pmid = 0)
  df <- rbind(df, data.frame(x = x, score = score))
}

df |>
  ggplot(aes(x = score, y = after_stat(density))) +
  geom_histogram(bins = 100, fill = "#2196F3") +
  labs(title = "Rating Distribution", x = "Score", y = "Density") +
  theme_minimal()
```

### Models

#### ZOIB Model

The Zero-One Inflated Beta (ZOIB) model assumes that the data can be modeled as a mixture of two logistic regression processes for the boundary values (0 and 1) and a beta regression process for the continuous proportions in-between.


```{r}
#| eval: false

f <- bf(
  score ~ x,
  phi ~ x,
  zoi ~ x,
  coi ~ x, 
  family = zero_one_inflated_beta()
)

m_zoib <- brm(f,
  data = df, family = zero_one_inflated_beta(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_zoib <- brms::add_criterion(m_zoib, "loo")  # For later model comparison

saveRDS(m_zoib, file = "models/m_zoib.rds")
```


#### XBX Model

[Kosmidis & Zeileis (2024)](https://arxiv.org/abs/2409.07233) introduce a generalization of the classic beta regression model with extended support [0, 1]. Specifically, the extended-support beta distribution (`xbeta`) leverages an underlying symmetric four-parameter beta distribution with exceedence parameter nu to obtain support [-nu, 1 + nu] that is subsequently censored to [0, 1] in order to obtain point masses at the boundary values 0 and 1.

```{r}
#| eval: false

f <- bf(
  score ~ x,
  phi ~ x,
  kappa ~ x, 
  family = xbeta()
)

m_xbx <- brm(f,
  data = df, family = xbeta(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_xbx <- brms::add_criterion(m_xbx, "loo")  # For later model comparison

saveRDS(m_xbx, file = "models/m_xbx.rds")
```


#### Beta-Gate Model

The [**Beta-Gate model**](https://dominiquemakowski.github.io/cogmod/reference/rbetagate.html) corresponds to a reparametrized Ordered Beta model ([Kubinec, 2023](https://doi.org/10.1017/pan.2022.20)). In this model, observed 0s and 1s represent instances where the underlying continuous response tendency fell beyond lower or upper boundary points ('gates').

```{r}
#| eval: false

f <- bf(
  score ~ x,
  phi ~ x,
  pex ~ x,
  bex ~ x, 
  family = betagate()
)

m_betagate <- brm(f,
  data = df, family = betagate(), stanvars = betagate_stanvars(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_betagate <- brms::add_criterion(m_betagate, "loo")  # For later model comparison

saveRDS(m_betagate, file = "models/m_betagate.rds")
```

#### CHOCO Model

See the [**documentation**](https://dominiquemakowski.github.io/cogmod/reference/rchoco.html) of the Choice-Confidence (CHOCO).

```{r}
#| eval: false

f <- bf(
  score ~ x,
  confright ~ x,
  confleft ~ x,
  precright ~ x,
  precleft ~ x,
  pex ~ x,
  bex ~ x,
  pmid = 0, 
  family = choco()
)

m_choco <- brm(f,
  data = df, family = choco(), stanvars = choco_stanvars(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_choco <- brms::add_criterion(m_choco, "loo")  # For later model comparison

saveRDS(m_choco, file = "models/m_choco.rds")
```


### Model Comparison

```{r}
#| echo: false

path <- "https://raw.github.com/DominiqueMakowski/cogmod/main/vignettes/models/"

m_zoib <- readRDS(url(paste0(path, "m_zoib.rds")))
m_xbx <- readRDS(url(paste0(path, "m_xbx.rds")))
m_betagate <- readRDS(url(paste0(path, "m_betagate.rds")))
m_choco <- readRDS(url(paste0(path, "m_choco.rds")))

df <- get_data(m_zoib)
```


#### Model Fit

We can compare these models together using the `loo` package, which shows
that CHOCO provides a significantly better fit than the other models.

```{r}
#| code-fold: true

# loo::loo_compare(m_zoib, m_xbx, m_betagate, m_choco) |> 
#   parameters(include_ENP = TRUE)
```


#### Sampling Duration

```{r}
rbind(
  data_modify(attributes(m_zoib$fit)$metadata$time$chain, Model="ZOIB"),
  data_modify(attributes(m_xbx$fit)$metadata$time$chain, Model="XBX"),
  data_modify(attributes(m_betagate$fit)$metadata$time$chain, Model="Beta-Gate"),
  data_modify(attributes(m_choco$fit)$metadata$time$chain, Model="CHOCO")
) |> 
  ggplot(aes(x = Model, y = total, fill = Model)) +
  geom_boxplot() +
  labs(y = "Sampling Duration (s)") +
  scale_fill_material_d(guide = "none") +
  scale_y_log10() +
  theme_minimal() 
```


#### Posterior Predictive Check

Running posterior predictive checks allows to visualize the predicted distributions from various models.
We can see how typical Beta-related models fail to capture the bimodal nature of the data, which is 
well captured by the CHOCO model.

Note: `iterations` controls the actual number of iterations used (e.g., for the point-estimate) and `keep_iterations` the number included.

```{r}
#| code-fold: true
#| eval: false

pred <- rbind(
  estimate_prediction(m_zoib, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "ZOIB"),
  estimate_prediction(m_xbx, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "XBX"),
  estimate_prediction(m_betagate, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "Beta-Gate"),
  estimate_prediction(m_choco, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "CHOCO")
)

df |>
  ggplot(aes(x = score, y = after_stat(density))) +
  geom_histogram(bins = 100, fill = "#2196F3") +
  labs(title = "Rating Distribution", x = "Score", y = "Density") +
  theme_minimal() + 
  geom_histogram(
    data = pred, aes(x = iter_value, group = as.factor(iter_group)),
    bins = 100, alpha = 0.02, position = "identity", fill = "#FF5722"
  ) +
  facet_wrap(~Model)
```


### Effect Visualisation

We can see how the predicted distribution changes as a function of **x** and gets "pushed" to the right. 
Moreover, we can also visualize the effect of **x** on specific parameters, showing that it mostly affects the parameter **conf** (the mean confidence - i.e., central tendency - on the right side), **confleft** (the relative confidence of the left side), and **mu**, which corresponds to the ***p*** probability of answering on the right. This is consistent with our expectations, and reflects 
the larger and more concentrated mass on the right of the scale for higher value of **x** (in brown).

```{r}
#| fig-width: 10
#| fig-height: 8
#| code-fold: true
#| eval: true

p1 <- estimate_prediction(m_choco, data = "grid", length = 4, keep_iterations = 500, iterations = 500) |> 
  reshape_iterations() |> 
  ggplot(aes(x = iter_value, fill = as.factor(x))) +
  geom_histogram(alpha = 0.6, bins = 100, position = "identity") +
  scale_fill_bluebrown_d() + 
  labs(x = "Rating") +
  theme_minimal()
p1

# # Predict various parameters
# pred_params <- data.frame()
# for(param in c("mu", "confright", "confleft", "precright", "precleft", "pex")) {
#   pred_params <- m_choco |>
#     estimate_prediction(data = "grid", length = 20, predict = param) |>
#     as.data.frame() |>
#     data_modify(Parameter = param) |> 
#     rbind(pred_params)
# }
# 
# p2 <- pred_params |>
#   ggplot(aes(x = x, y = Predicted)) +
#   geom_ribbon(aes(ymin = CI_low, ymax = CI_high, fill = Parameter), alpha = 0.2) +
#   geom_line(aes(color = Parameter), linewidth = 1) +
#   facet_wrap(~Parameter, scales = "free_y", ncol=3) +
#   scale_fill_viridis_d() +
#   scale_color_viridis_d() +
#   theme_minimal()
# p2
```


```{r}
#| echo: false
#| eval: false

library(patchwork)

ggsave("../man/figures/subjective_ratings1.png", p1 / p2, width = 12, height = 8, bg = "white")
```

## Likert Scales

### Simulate Data

```{r}
#| fig-width: 10
#| fig-height: 7
#| warning: false
#| message: false

set.seed(33)

df <- data.frame()
for(x in seq(0.2, 0.8, by = 0.1)) {
  rating <- rbetadiscrete(n=200, mu = x, phi = 3, k = 6)
  df <- rbind(df, data.frame(x = x, rating = rating))
}

df |>
  ggplot(aes(x = rating)) +
  geom_bar(stat = "count", fill = "#2196F3") +
  labs(title = "Rating Distribution", x = "Score", y = "Density") +
  theme_minimal()
```

### Models


#### Beta-Binomial

The Beta-Binomial is a common default for bounded count data, though it wasn't designed for it. Its distribution consists of a two-stage mixture: conditional on a success probability *p*, the response follows a $Binomial(n, p)$, while *p* itself varies according to a Beta distribution. This lets it absorb extra-binomial variation (overdispersion) arising when raters differ systematically in their response tendencies. This model makes sense when a rating is literally a count of successes a(e.g. items endorsed out of a checklist), but for a single holistic judgment on a *K*-point scale, it's better understood as a descriptive tool than a model of the response process itself.

**NOTE:** the Beta-Binomial distribution in brms is defined on [0, n], so we need to shift the ratings down by 1 (1-6 -> 0-5).

```{r}
#| eval: false

f <- bf(
  rating | trials(5) ~ x,
  phi ~ x,
  family = brms::beta_binomial()
)

m_betabinomial <- brm(f,
  data = mutate(df, rating = rating - 1), 
  family = brms::beta_binomial(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_betabinomial <- brms::add_criterion(m_betabinomial, "loo")  # For later model comparison

saveRDS(m_betabinomial, file = "models/m_betabinomial.rds")
```


#### Beta-Discrete

The Discrete Beta model ([Sciandra, 2024](https://link.springer.com/article/10.1007/s10651-023-00592-5)) was developed for modeling discrete rating data, i.e., discrete responses on a fixed integer scale (1-K). It assumes the observed rating is the discretised counterpart of a continuous, unobservable latent response, but rather than estimating $K-1$ threshold parameters (like the Ordinal model), the Discrete Beta fixes these thresholds at evenly spaced points, and instead lets a Beta distribution govern the proportions. This reverses the usual logic (fixed distribution, estimated thresholds) and, because the Beta distribution can take on a wide range of shapes (including non-monotonic convex shapes like "J" and "U" shapes that many ordinal models struggle to reproduce), it yields a distribution that is both highly flexible and remarkably parsimonious, requiring only two parameters regardless of how many categories the scale has.


```{r}
#| eval: false

f <- bf(
  rating | vint(6) ~ x,
  phi ~ x,
  family = betadiscrete()
)

m_betadiscrete <- brm(f,
  data = df, family = betadiscrete(), stanvars = betadiscrete_stanvars(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_betadiscrete <- brms::add_criterion(m_betadiscrete, "loo")  # For later model comparison

saveRDS(m_betadiscrete, file = "models/m_betadiscrete.rds")
```

#### Ordinal (Cumulative)

Ordinal models (also known as cumulative models) posit a continuous latent variable, partitioned by $K-1$ estimated thresholds. Because thresholds are freely estimated rather than fixed, the model doesn't assume equally-spaced categories. This maps onto signal-detection accounts of ordinal judgment, where a continuous perception is filtered through response criteria that can be unevenly spaced (e.g. reluctance to use extremes). That flexibility comes at the cost of extra parameters relative to Beta-Discrete, and these thresholds parameters are estimated on an arbitrary latent scale, which are not really interpretable in and of themselves. 


```{r}
#| eval: false

f <- bf(
  rating ~ x,
  disc ~ x,
  family = brms::cumulative()
)

m_cumulative <- brm(f,
  data = df, family = brms::cumulative(), init = 0,
  chains = 4, iter = 500, backend = "cmdstanr"
)

m_cumulative <- brms::add_criterion(m_cumulative, "loo")  # For later model comparison

saveRDS(m_cumulative, file = "models/m_cumulative.rds")
```


### Model Comparison

```{r}
#| echo: false

path <- "https://raw.github.com/DominiqueMakowski/cogmod/main/vignettes/models/"

m_betabinomial <- readRDS(url(paste0(path, "m_betabinomial.rds")))
m_betadiscrete <- readRDS(url(paste0(path, "m_betadiscrete.rds")))
m_cumulative <- readRDS(url(paste0(path, "m_cumulative.rds")))

df <- get_data(m_betadiscrete)
```


We can compare these models together using the `loo` package.
```{r}
#| code-fold: true
#| eval: false

loo::loo_compare(m_betabinomial, m_betadiscrete, m_cumulative) |>
  report::report()
```



```{r}
#| code-fold: true
#| eval: false

pred <- rbind(
  estimate_prediction(m_betabinomial, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(iter_value = iter_value + 1, Model = "Beta-Binomial"),
  estimate_prediction(m_betadiscrete, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "Beta-Discrete"),
  estimate_prediction(m_cumulative, keep_iterations = 50, iterations = 50) |>
    reshape_iterations() |>
    data_modify(Model = "Cumulative")
)  
  
  
  
pred_counts <- pred |> 
  summarize(n = n(), .by = c("Model", "iter_group", "iter_value")) |> 
  summarize(n_median = median(n),
            ci_low = quantile(n, 0.025),
            ci_high = quantile(n, 0.975),
            .by = c("Model", "iter_value"))


df |>
  ggplot(aes(x = rating)) +
  geom_bar(stat = "count", fill = "#2196F3") +
  geom_line(data = pred_counts, aes(x = iter_value, y = n_median), color = "#FF5722", linewidth = 1) +
  geom_pointrange(
    data = pred_counts,
    aes(x = iter_value, y = n_median, ymin = ci_low, ymax = ci_high),
    color = "#FF5722", size = 1
  ) +
  labs(title = "Rating Distribution", x = "Score", y = "Density") +
  theme_minimal() +
  facet_wrap(~Model)
```
