---
title: "Outliers and Non-Decision Time"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
format:
  html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(cogmod)
library(ggplot2)

theme_set(theme_minimal())
```

```{r}
#| label: load-caches
#| echo: false

path <- "https://raw.github.com/DominiqueMakowski/cogmod/main/vignettes/models/"

desc <- readRDS(url(paste0(path, "outliers_desc.rds")))
hst <- readRDS(url(paste0(path, "outliers_hist.rds")))
caf <- readRDS(url(paste0(path, "outliers_caf.rds")))
perds <- readRDS(url(paste0(path, "outliers_perdataset.rds")))
sens <- readRDS(url(paste0(path, "outliers_sensitivity.rds")))
rec <- readRDS(url(paste0(path, "outliers_recovery.rds")))
```

## The Problem with Non-Decision Time

Every shifted reaction time distribution has a non-decision time `ndt`: the
incompressible encoding and motor delay before evidence accumulation begins.
And every shifted distribution assigns **exactly zero density** to any response
faster than `ndt`. That is not a modelling choice, it follows from the
definition: `RT = ndt + decision_time`, and decision time cannot be negative.

The consequence is a hard boundary in the likelihood. If a single response in
the dataset is faster than the `ndt` a sampler is currently considering, the
log-density is `-Inf` and the whole posterior collapses. So `ndt` is bounded
above by the *fastest observed response*, which is an order statistic of the
sample rather than a property of the participant.

Earlier versions of `cogmod` handled this by injecting `min(RT)` as data and
expressing the shift as a proportion of it, `ndt = tau * minrt` with
`tau` in `(0, 1)`. That sampled well, but it capped `ndt` at the fastest
response *in the entire dataset*. As soon as `ndt` varied by condition or by
participant, the cap was shared across all of them, and any group whose true
non-decision time exceeded the global minimum could not be represented at all.

`cogmod` now estimates `ndt` directly, in seconds, and makes that possible by
mixing a small proportion `poutlier` of responses from an **outlier
distribution** that has support over the whole positive line:

$$
p(\text{RT}) = (1 - \text{poutlier}) \cdot f(\text{RT} - \text{ndt}) +
  \text{poutlier} \cdot g(\text{RT})
$$

Below `ndt` the first term is still exactly zero, but the second is positive.
The boundary becomes a finite *cost* rather than a wall, the log-density stays
smooth and differentiable, and nothing needs to be taken from the data.

The component `g` is hard-coded as `LogNormal(log(0.15), 1.5)`:

```{r}
#| label: fig-component
#| fig-cap: "*The hard-coded outlier component: LogNormal(log(0.15), 1.5). Median 150 ms; the dashed lines mark the 2.5th and 97.5th percentiles.*"
#| fig-height: 3

q <- qlnorm(c(0.025, 0.975), log(0.15), 1.5)

ggplot() +
  geom_function(fun = dlnorm, args = list(meanlog = log(0.15), sdlog = 1.5),
                n = 1000, xlim = c(0.001, 3), linewidth = 0.8) +
  geom_vline(xintercept = q, linetype = "dashed", colour = "grey50") +
  labs(x = "RT (s)", y = "density")
```

It is deliberately broad - 95% of its mass falls between
`r round(q[1], 3)` s and `r round(q[2], 2)` s - because an outlier is not
necessarily an *extreme* value. A response driven by some process other than
the decision can land anywhere in the plausible range. But it peaks in the fast
region, where the decision component has no density to offer.

This article asks whether that specification is defensible.

## The Data

Validating this needs reaction times that have **not** been trimmed, which is
unusual: almost every published dataset has had its fast responses removed
before distribution. The [Illusion Game](https://github.com/RealityBending/IllusionGameComputational)
data are untrimmed, very large, and - crucially - include accuracy, which will
let us identify contaminant trials without using an RT model at all.

```{r}
#| eval: false
#| code-fold: false

repo <- paste0("https://raw.githubusercontent.com/RealityBending/",
               "IllusionGameComputational/refs/heads/main/data/")

df <- rbind(
  read.csv(paste0(repo, "illusion_part1.csv")),
  read.csv(paste0(repo, "illusion_part2.csv")),
  read.csv(paste0(repo, "illusion_part3.csv"))
)
```

```{r}
#| echo: false

cat(sprintf("%s trials, %s participants, RT from %.3f to %.2f s\n",
            format(desc$n, big.mark = ","), format(desc$n_ppt, big.mark = ","),
            desc$rt_range[1], desc$rt_range[2]))
cat(sprintf("overall error rate: %.3f\n", desc$err))
cat(sprintf("responses faster than 0.30 s: %d (%.3f%%)\n",
            desc$n_fast, desc$pct_fast))
```

Participants judged which of two shapes was larger or longer under various
visual illusions - a two-alternative discrimination, so **chance performance is
an error rate of 0.5**. The fastest recorded response is 1 ms, which is not a
decision about anything.

To keep the analyses below quick, and because the whole dataset is a
million trials, the model fits use a single sub-study
(`r format(attr(sens, "n"), big.mark = ",")` trials from
`r attr(sens, "n_ppt")` participants). The descriptive analyses use everything.

## Why the Old Parameterization Cannot Work Here

The fastest response in the modelling subset is
**`r sprintf("%.3f", attr(sens, "min_rt"))` s**. Under `ndt = tau * minrt`,
that would have been the ceiling on non-decision time for every participant and
every condition - a physiological impossibility, and roughly two orders of
magnitude below any plausible value.

This is not a subtle bias. It is what the fully un-trimmed case looks like, and
it is the reason the parameterization changed.

## Model-Free Evidence: the Conditional Accuracy Function

Before fitting anything, accuracy alone can tell us whether fast contamination
is real and where it lives. A genuine decision, however hurried, should perform
above chance. A contaminant response - a key press with no decision behind it -
should be at chance regardless of how fast it was.

```{r}
#| label: fig-caf
#| fig-cap: "*Error rate as a function of RT. The dashed line is chance (0.5) for this two-alternative task; the dotted line is the error rate of unambiguously genuine decisions. Point size is proportional to the number of trials in the bin.*"
#| fig-height: 3.6

err_good <- with(caf, sum(err * n * (lo >= 0.4 & hi <= 1.5)) /
                   sum(n * (lo >= 0.4 & hi <= 1.5)))

ggplot(caf[caf$mid < 1.5, ], aes(x = mid, y = err)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "red") +
  geom_hline(yintercept = err_good, linetype = "dotted") +
  geom_line(colour = "grey40") +
  geom_point(aes(size = n), alpha = 0.7) +
  scale_size_area(max_size = 5, guide = "none") +
  labs(x = "RT (s)", y = "error rate")
```

```{r}
#| echo: false

k <- caf[caf$hi <= 0.45, c("lo", "hi", "n", "err")]
k$err <- round(k$err, 3)
knitr::kable(k, col.names = c("RT from", "RT to", "n", "error rate"))
```

The result is unambiguous. Below 0.30 s the error rate sits flat at chance,
then climbs steeply to the genuine-decision baseline of
`r sprintf("%.3f", err_good)` by about 0.45 s. That is direct evidence, owing
nothing to any RT model, that

- fast contamination is **real** and not an artefact of distributional
  assumptions,
- it occupies roughly **[0, 0.30] s** in this task,
- it accounts for about **`r sprintf("%.1f%%", desc$pct_fast)`** of trials here.

This also justifies the mixture *form* independently of the hyperparameters. A
model that assigns zero density to these trials is not describing them
conservatively, it is declaring them impossible.

## What Accuracy Cannot Tell Us: the Slow Tail

The same logic tempts one to extend the analysis upward, and it fails. Notice
in the figure that the error rate *rises again* for slow responses, from
`r sprintf("%.3f", min(caf$err))` around 0.5 s to
`r sprintf("%.3f", caf$err[nrow(caf)])` in the slowest bin. Naively treating
that excess as contamination implies over 7% contaminant trials with a median
near 1 s.

That inference is wrong. Slow trials are *hard* trials - strong illusions and
small differences - so difficulty confounds RT and accuracy at the slow end. A
slow contaminant is statistically indistinguishable from the right tail of the
RT distribution itself, which is precisely why `cogmod` does not attempt to
model one. Implausibly slow responses should be filtered before fitting; unlike
fast responses, the model cannot identify them for you.

## Does the Hard-Coded Choice Matter?

The hyperparameters of `g` are not a claim about how fast guesses are
distributed. Their job is narrower: to put positive density below `ndt` so that
the boundary dissolves. The right test is therefore not "are these the true
values?" but "does the answer depend on them?"

Refitting the same data under seven very different outlier components:

```{r}
#| echo: false

s <- sens
s$ndt <- round(s$ndt, 4)
s$poutlier <- round(s$poutlier, 4)
s$logLik <- ifelse(s$logLik < -1e8, NA, round(s$logLik, 1))
knitr::kable(s, col.names = c("outlier component g", "ndt (s)", "poutlier",
                              "logLik"))
```

```{r}
#| echo: false

ok <- sens[sens$contaminant != "none (poutlier = 0)", ]
cat(sprintf("ndt across all seven specifications: %.3f to %.3f s (a %.0f ms range)\n",
            min(ok$ndt), max(ok$ndt), 1000 * diff(range(ok$ndt))))
```

Two things stand out.

**`ndt` barely moves.** Components that differ by orders of magnitude in shape,
location and boundedness - including a uniform spanning the entire RT range -
all recover a non-decision time within about 15 ms of each other. The parameter
of scientific interest is effectively invariant to the choice, which is exactly
what a regularizing device should deliver.

**`poutlier` does move**, from
`r sprintf("%.3f", min(ok$poutlier))` to `r sprintf("%.3f", max(ok$poutlier))`.
It is a rate defined *relative to an assumed reference distribution*, not an
absolute proportion of bad trials, so it is not comparable across differently
specified models. Report the component alongside it.

**Without any outlier component the model cannot be fitted at all.** The last
row of the table is not a poor estimate; it is a failure. With a fastest
response of `r sprintf("%.3f", attr(sens, "min_rt"))` s the likelihood has no
usable region, and the optimizer never leaves its starting value.

## The Shape Is Not Optimal, and That Is Informative

Being honest about it: the shipped component is *not* the best-fitting one. The
empirical distribution of the chance-level responses is roughly flat from 0 to
about 0.19 s and then rises steeply toward 0.30 s, whereas
`LogNormal(log(0.15), 1.5)` decreases monotonically across that range.

```{r}
#| label: fig-fasttail
#| fig-cap: "*Observed responses in 10 ms bins (bars) against the hard-coded outlier component (line, rescaled to the counts). The component decreases where the data are flat and then rising.*"
#| fig-height: 3.4

fast <- hst[hst$mid < 0.42, ]
scl <- max(fast$count[fast$mid < 0.2]) / dlnorm(0.1, log(0.15), 1.5)

ggplot(fast, aes(x = mid, y = count)) +
  geom_col(width = 0.0095, fill = "grey65") +
  geom_line(aes(y = dlnorm(mid, log(0.15), 1.5) * scl),
            colour = "red", linewidth = 0.8) +
  coord_cartesian(ylim = c(0, max(fast$count[fast$mid < 0.32]) * 1.1)) +
  labs(x = "RT (s)", y = "number of trials")
```

There is a sensible reading. A contaminant response is a motor act with no
decision behind it, and simple reaction time peaks around 0.2-0.3 s. True
anticipations - pressing before the stimulus registers at all - sit near zero
and are rare. So the fast tail is a thin flat anticipation floor plus a
guessing component at simple-RT latency, and a distribution peaked at 0.15 s
splits the difference between the two.

Consistently with that, the table above shows that moving the peak *later*
improves the fit (`LogNormal(log(0.25), 1.0)` gains roughly 300 log units) while
moving it *earlier* makes things worse.

The component was nonetheless left as it is, for two reasons. `ndt` is
invariant, so the better-fitting shapes buy no accuracy in the quantity anyone
reports. And a narrower component puts less density at the extreme fast end,
which is where the boundary protection has to work for tasks whose `ndt` is far
below the 0.32 s seen here. `LogNormal(log(0.25), 1.0)` is recorded here as the
evidence-based candidate should the default ever be revisited.

::: {.callout-note}
One caveat on generality: this is one task, in one modality, with a particular
response device. The flat-then-rising shape may not transfer. What should
transfer is the invariance of `ndt`, which is a property of the mixture rather
than of the dataset.
:::

## Parameter Recovery

The insensitivity above says the estimate is stable, not that it is correct.
For that, simulate from the model with a known non-decision time using
`rrt_lognormal()` and fit it back.

```{r}
#| eval: false
#| code-fold: false

x <- rrt_lognormal(500, mu = -0.7, sigma = 0.5, ndt = 0.35, poutlier = 0.02)
```

```{r}
#| echo: false

cat(sprintf("40 replications, n = 500 per replication\n"))
cat(sprintf("  ndt      : mean %.4f (sd %.4f)   truth 0.350\n",
            mean(rec$ndt), sd(rec$ndt)))
cat(sprintf("  poutlier : mean %.4f (sd %.4f)   truth 0.020\n",
            mean(rec$poutlier), sd(rec$poutlier)))
```

Both parameters recover, and `poutlier` recovers well when the assumed
component matches the generating one - which is the case here by construction,
and is the best case rather than the typical one.

## Per-Trial Outlier Probabilities

`poutlier` is a rate, not a classification: the model never labels individual
trials. But the mixture does assign each response a posterior probability of
having come from the outlier component, which `contaminant_prob()` returns:

```{r}
#| eval: false
#| code-fold: false

f <- brms::bf(RT ~ 1, sigma ~ 1, ndt ~ 1, poutlier ~ 1,
              family = rt_lognormal())

m <- brms::brm(f, data = df, stanvars = rt_lognormal_stanvars())

head(contaminant_prob(m))
```

The quantity is
`poutlier * g(rt) / (poutlier * g(rt) + (1 - poutlier) * f(rt - ndt))`.
Responses faster than `ndt` come out at 1, responses in the heart of the
distribution near 0, and responses in either tail somewhere in between. The
grading is driven by evidence rather than by a cutoff, and it lines up with the
conditional accuracy function without ever having been shown the accuracy data.

The interpretation needs care in one respect. A low probability does **not**
mean a trial is clean; it means the data cannot tell. A lapse that happens to
produce a plausible reaction time is statistically invisible, and the mixture
handles it correctly by accounting for its *rate* without pretending to
identify it. No trimming rule can do that - a cutoff can only remove what looks
extreme, which is both the wrong trials and not enough of them.

## A Note on Reading Density Plots of the Fast Tail

One practical warning, because it is easy to be misled. Plotting the fast tail
split by sub-study suggests a local peak near 0.1 s that is not in the pooled
data. The reason is sample size:

```{r}
#| echo: false

pd <- perds[order(perds$n_fast), ]
knitr::kable(pd, row.names = FALSE,
             col.names = c("Dataset", "n trials", "n below 0.25 s"))
```

Some sub-studies contribute barely a dozen responses to the whole region. Since
a kernel density is renormalized *within* each group, those few trials are
stretched into a full-height curve, and one or two of them become a "peak".
Pooled, the shape is flat and then rising, with no local maximum near 0.1 s.
When inspecting a region that holds a fraction of a percent of the data, plot
counts rather than densities.

## Summary

- A shifted RT distribution gives zero density below `ndt`, which bounds
  non-decision time by the fastest observed response. On untrimmed data that
  makes the unmixed model unfittable, not merely biased.
- Mixing in a broad outlier component removes the boundary while keeping the
  log-density smooth, and lets `ndt` be estimated directly in seconds.
- Accuracy provides model-free confirmation: responses below 0.30 s in this task
  are at chance, and constitute about half a percent of trials.
- The hard-coded hyperparameters of that component are a regularizing choice,
  not an empirical claim. `ndt` moves by ~15 ms across wildly different
  specifications, which is the property that matters.
- The shipped shape is not the best-fitting one; `LogNormal(log(0.25), 1.0)`
  fits this dataset better and is noted as a candidate for future revision.
- Slow contamination is *not* handled by the mixture, because it is confounded
  with the right tail of the RT distribution. Filter it beforehand.
