---
title: "Outliers and Non-Decision Time"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
format:
  html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(cogmod)
library(ggplot2)
library(dplyr)
```

Every object used below is built by code shown in the article. Because the data
are large, the results are cached; each section shows the code that produced its
own piece, and the final chunk writes the cache. To rebuild from scratch, run the
`eval: false` chunks in order.

```{r}
#| label: load-cache
#| echo: false

path <- "https://raw.github.com/DominiqueMakowski/cogmod/main/vignettes/models/"
cache <- readRDS(url(paste0(path, "outliers.rds")))

desc      <- cache$desc
hist_fast <- cache$hist
caf       <- cache$caf
baseline  <- cache$baseline
sens      <- cache$sensitivity
fitinfo   <- cache$fitinfo
rates     <- cache$rates
```

## The Problem of Non-Decision Time

A shifted reaction time model (e.g., Shifted LogNormal or any Evidence
Accumulation Model) assigns **exactly zero density** to any response faster than
its non-decision time `ndt`. The consequence is a hard boundary: if one response
in the dataset is faster than the `ndt` a sampler is considering, the log-density
is `-Inf` and the posterior collapses. `ndt` is therefore bounded above by the
fastest observed response: a constant taken from the sample rather than a
property of the participant or the condition. Worse, that bound is *shared*: one
anomalously fast trial caps `ndt` for every condition and every participant in
the model.

This is not specific to `cogmod` - every implementation of a shifted RT model has
to deal with it. HDDM estimates the
shift directly under a bounded prior and ships a `p_outlier` parameter with a
uniform contaminant - the same mixture idea used below. HSSM estimates it over
the range its likelihood-approximation networks were trained on. EMC2 and DMC put
the shift on a log scale and lean on particle-based samplers, which tolerate a
hard boundary by rejecting proposals that cross it. Gradient-based samplers cannot afford a cliff, 
so the mixture is what makes the problem tractable for HMC.

In `cogmod`, the boundary is removed by mixing the RT distribution with a
proportion of responses from an **outlier component**.

$$
p(\text{RT}) = (1 - \text{poutlier}) \cdot f(\text{RT} - \text{ndt}) + \text{poutlier} \cdot g(\text{RT})
$$

Below `ndt` the first term is still zero, but the second is positive. The
boundary becomes a finite *cost* rather than a wall, the log-density stays smooth
and differentiable, and `ndt` can be estimated directly in seconds with nothing
taken from the data.

## Choosing the Outlier Component

`g` is a **half Normal** with scale `0.2` seconds, and that is what every
number below uses. Four properties decide the shape, and every obvious
candidate fails at least one.

```{r}
#| code-fold: true
#| code-summary: "Analytic properties of each candidate"

hn <- function(s) function(x) 2 * dnorm(x, 0, s)
ht <- function(s, nu) function(x) 2 * dt(x / s, nu) / s

candidates <- list(
  "Half-Normal(0.2)  [shipped]" = hn(0.2),
  "Half-t(0.3, df = 3)  [<= 0.2.0]" = ht(0.3, 3),
  "Half-Cauchy(0.3)"               = ht(0.3, 1),
  "LogNormal(log 0.15, 1.5)"       = function(x) dlnorm(x, log(0.15), 1.5),
  "Exponential(mean 0.15)"         = function(x) dexp(x, 1 / 0.15),
  "Uniform(0, 0.35)"               = function(x) ifelse(x <= 0.35, 1 / 0.35, 0),
  "Uniform(0, 2)"                  = function(x) ifelse(x <= 2, 1 / 2, 0)
)

props <- do.call(rbind, lapply(names(candidates), function(nm) {
  g <- candidates[[nm]]
  m <- try(integrate(function(z) z * g(z), 0, Inf,
                     subdivisions = 4000)$value, silent = TRUE)
  data.frame(
    g = nm,
    # density relative to its value near zero: 1.00 is perfectly flat
    flatness = sprintf("%.2f / %.2f / %.2f",
                       g(0.05) / g(1e-4), g(0.15) / g(1e-4), g(0.25) / g(1e-4)),
    tail = format(g(3) / g(1e-4), digits = 2),
    mean = if (inherits(m, "try-error") || m > 10) "divergent" else sprintf("%.3f", m)
  )
}))

knitr::kable(props, col.names = c("g", "flatness at 0.05 / 0.15 / 0.25 s",
                                  "g(3 s), relative to peak", "mean"))
```

- **Flat at the origin.** The fastest responses are the ones least plausibly
decisions, so `g` must not thin out there. The flatness column gives the density
relative to its value near zero, so `1.00` is perfectly flat. A LogNormal rises
and collapses toward zero exactly where protection is needed most. An Exponential
has the opposite defect, peaking at zero with maximal slope. Both encode a claim
nobody would defend: that anticipations are far more likely at e.g., 150 ms than
at 3 ms. The half-Normal, half-t and uniform are all genuinely flat.
- **Dead by the time real decisions start.** This is the criterion the half-t
fails, and it is why it was replaced in 0.2.1. Its tails are heavier than every
decision density in the package, so far-out slow responses end up better
explained by `g` than by the model itself: at `poutlier = 0.02`, against a
shifted LogNormal, a 5 s response was attributed to the outlier component with
probability 0.86, and the crossover sat at 3.86 s. `ndt` then gets pulled up
behind it. The half-Cauchy and the wide uniform fail this worse. The slow tail
belongs to the decision family - that is what `shape` in `cogmod_loggamma()`
and `sigmadrift` in `cogmod_invgaussian()` are for.
- **Strictly positive everywhere.** The whole point of mixing `g` in is that
no response should ever have zero density; a single one that does makes the
log-likelihood `-Inf` and puts back a cliff of exactly the kind being removed.
That is what rules out the uniforms, which are otherwise flat, tail-free and
well behaved: put the edge at 0.35 s and a contaminant at 0.36 s under an `ndt`
of 0.40 s is declared impossible. A half-Normal is never zero - merely
negligible, which is a different thing.
- **Finite mean.** `posterior_epred()` needs `E[RT]`. A half-Cauchy has no finite
mean, so it would return infinity.

The half-Normal is the only candidate that satisfies all four, and it does so
without a trade: because `exp(-x^2 / 2s^2)` annihilates the far tail at *any*
scale, flatness near zero and tail weight are not bought against each other the
way they are for a Student-t. At 0.2 s it holds 76% of its peak density at
0.15 s and 46% at 0.25 s - against 85% and 66% for the half-t it replaced -
while being 16 orders of magnitude lighter at 3 s.

```{r}
#| label: fig-component
#| fig-cap: "*The outlier component: half-Normal(0, 0.2), flat near the origin and gone before real decisions begin.*"
#| fig-height: 3

ggplot() +
  geom_function(fun = function(x) 2 * dnorm(x, 0, 0.2),
                n = 1000, xlim = c(0.0001, 3), linewidth = 0.8) +
  labs(x = "RT (s)", y = "density")
```

## Reaction Times Must Be in Seconds

The scale is `0.2` s, a **constant**, and it is not configurable. Whatever value
it took, the component's job is to stay flat across the range `ndt` plausibly
occupies - the `ndt` prior puts that at 0.20 to 0.45 s - and then to disappear.
0.2 s does that: 68% of its mass falls below 0.2 s and 92% below 0.35 s, and a
contaminant landing just under a late `ndt` of 0.42 s still has a log-density of
-4.6, where the half-t gave -4.0. Nothing that used to be covered is starved.

Up to 0.2.0 the scale was a `minrt` argument in the unit of the data, which made
the likelihood **equivariant** to that unit. That argument was removed in 0.2.1,
because the equivariance was already fictional end to end: `cogmod_priors()`
shifted only the `ndt` prior with it, while the `sigmandt` prior of
`cogmod_ddm()`, the `sigmadrift` prior of `cogmod_invgaussian()` and the `mu`
priors are all stated in seconds outright - and `cogmod_priors()` is not
optional. The package works in seconds; the constant now says so, and one
argument disappears from every family, every density and every `stanvars()`.

Feeding it milliseconds fails **silently**, which is worth knowing about. The
outlier component's log-density at `RT = 400` is about `-2e6`, so it contributes
nothing anywhere in the data and the mixture collapses back to the unmixed
shifted family: `poutlier` goes to zero and `ndt` is pinned by the fastest
observed response again - exactly the boundary this parameterization exists to
remove. Nothing errors, and the chains still initialise, because the decision
density itself stays finite. Divide by 1000 before fitting.

## The Data

Validating this needs reaction times that have **not** been trimmed, which is
unusual - almost every published dataset has its fast responses removed before
distribution. The [Illusion Game](https://github.com/RealityBending/IllusionGameComputational)
data are untrimmed, large, and include accuracy, which will let us identify
contaminant trials without an RT model at all.

```{r}
#| eval: false
#| code-fold: true

repo <- paste0("https://raw.githubusercontent.com/RealityBending/",
               "IllusionGameComputational/refs/heads/main/data/")

df <- rbind(
  read.csv(paste0(repo, "illusion_part1.csv")),
  read.csv(paste0(repo, "illusion_part2.csv")),
  read.csv(paste0(repo, "illusion_part3.csv"))
)

desc <- list(
  n = nrow(df), n_ppt = length(unique(df$Participant)),
  rt_range = range(df$RT), err = mean(df$Error),
  n_fast = sum(df$RT < 0.30), pct_fast = 100 * mean(df$RT < 0.30)
)
```

```{r}
#| echo: false

cat(sprintf("%s trials, %s participants, RT from %.3f to %.2f s\n",
            format(desc$n, big.mark = ","), format(desc$n_ppt, big.mark = ","),
            desc$rt_range[1], desc$rt_range[2]))
cat(sprintf("responses faster than 0.30 s: %d (%.3f%%)\n",
            desc$n_fast, desc$pct_fast))
```

Participants judged which of two shapes was larger or longer under various
visual illusions - a two-alternative discrimination, so **chance performance is
an error rate of 0.5**. The fastest recorded response is 1 ms.

Model fits below use one sub-study
(`r format(fitinfo$n, big.mark = ",")` trials, `r fitinfo$n_ppt` participants)
for speed; descriptives use everything.

## Model-Free Evidence

Accuracy alone establishes whether fast contamination is real, without fitting
anything. A genuine decision, however hurried, should perform above chance. A
key press with no decision behind it should be at chance however fast it was.

To see whether the pattern is task-general, the same curve is drawn for two very
different paradigms, both shipped with `rtdists` and both taken **uncensored** -
including the trials the original authors flagged for removal, since those are
precisely the ones of interest.

The first is the lexical decision data of Wagenmakers et al. (2008),
`speed_acc`, split by its speed/accuracy instruction. The second is the
brightness discrimination of Ratcliff and Rouder (1998), `rr98`: a perceptual
two-choice task with no response conflict, which matters because in a flanker or
Stroop task - and, to a lesser extent, in an illusion judgement - fast errors can
be genuine responses driven by a competing stimulus dimension rather than
contamination. `rr98` has no such competing dimension, and it ships raw: its
`outlier` column *flags* responses below 0.2 s or above 2.5 s instead of deleting
them.

```{r}
#| eval: false
#| code-fold: true

caf_of <- function(rt, err, brk, series) {
  b <- cut(rt, brk, include.lowest = TRUE)
  ag <- data.frame(e = err, b = b) |>
    summarize(n = length(e), m = mean(e), .by = b)
  i <- match(ag$b, levels(b))
  data.frame(lo = brk[i], hi = brk[i + 1], n = ag$n,
             err = ag$m, series = series)
}

# Illusion Game: dense fast tail, so fine bins are affordable
caf <- caf_of(df$RT, df$Error,
              c(seq(0, 0.50, by = 0.025), 0.6, 0.8, 1.0, 1.2, 1.5, 6),
              "Illusion Game (all)")

# Lexical decision of Wagenmakers (2008), ALL trials. `response == "error"` marks uninterpretable
# trials, for which accuracy is undefined. Its fast tail is far sparser, so
# coarser bins are needed.
data(speed_acc, package = "rtdists")
sa <- speed_acc[speed_acc$response != "error", ]
sa$Error <- as.integer(as.character(sa$response) != as.character(sa$stim_cat))
brk_s <- c(seq(0, 0.50, by = 0.05), 0.6, 0.8, 1.0, 1.2, 1.5, 3, 1200)

for (cd in levels(sa$condition)) {
  s <- sa[sa$condition == cd, ]
  caf <- rbind(caf, caf_of(s$rt, s$Error, brk_s,
                           paste0("Wagenmakers2008 (", cd, ")")))
}

# Brightness discrimination of Ratcliff & Rouder (1998), ALL trials: the
# `outlier` column flags rt < 0.2 or rt > 2.5 rather than removing them, so the
# untrimmed data are already here. Pooled over the speed/accuracy instruction
# for the figure; the split is quantified in the next section.
data(rr98, package = "rtdists")
caf <- rbind(caf, caf_of(rr98$rt, as.integer(!rr98$correct), brk_s,
                         "RatcliffRouder1998 (all)"))

caf$mid <- (caf$lo + caf$hi) / 2
caf$series <- factor(caf$series, levels = c(
  "Illusion Game (all)", "Wagenmakers2008 (speed)",
  "Wagenmakers2008 (accuracy)", "RatcliffRouder1998 (all)"))

# Error rate of unambiguously genuine decisions, per series
baseline <- do.call(rbind, lapply(levels(caf$series), function(sr) {
  d <- caf[caf$series == sr & caf$lo >= 0.4 & caf$hi <= 1.5, ]
  data.frame(series = sr, err_good = sum(d$err * d$n) / sum(d$n))
}))
baseline$series <- factor(baseline$series, levels = levels(caf$series))
```

The outlier component is shaded behind, normalized so its maximum sits at chance.

```{r}
#| label: fig-caf
#| fig-cap: "*Error rate by RT across three paradigms. Dashed red line is chance (0.5); dotted lines are each series' own error rate for unambiguously genuine decisions. Point size is proportional to trials per bin; bins with fewer than 5 trials are dropped. The grey area is the half-Normal outlier component, rescaled to the panel (arbitrary units).*"
#| fig-height: 4

dens <- data.frame(x = seq(0.001, 2, length.out = 600))
dens$y <- (2 * dnorm(dens$x, 0, 0.2)) / (2 * dnorm(0, 0, 0.2)) * 0.5

ggplot(caf[caf$mid < 2 & caf$n >= 5, ], aes(x = mid, y = err, colour = series)) +
  geom_area(data = dens, aes(x = x, y = y), inherit.aes = FALSE,
            fill = "grey75", alpha = 0.45) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "red") +
  geom_hline(data = baseline, aes(yintercept = err_good, colour = series),
             linetype = "dotted", linewidth = 0.4, show.legend = FALSE) +
  geom_line(linewidth = 0.6) +
  geom_point(aes(size = n), alpha = 0.85) +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 1, 1.5), expand = expansion(add = c(0, 0.1))) +
  scale_size_area(max_size = 5, guide = "none") +
  scale_colour_manual(values = c("grey20", "#D55E00", "#0072B2", "#009E73")) +
  labs(x = "RT (s)", y = "error rate", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "top")  +
  coord_cartesian(xlim = c(0, 1.5))
```

Below 0.30 s the error rate sits at chance, then climbs to the genuine-decision
baseline by about 0.45 s. Fast contamination is real, it occupies roughly
`[0, 0.30]` s, and it accounts for `r sprintf("%.1f%%", desc$pct_fast)` of trials
in the Illusion Game data. A model that assigns these responses zero density is
not being conservative; it is calling them impossible.

Three features of the comparison matter.

The **plateau replicates across paradigms** while the baseline does not. The four
series settle at
`r paste(sprintf("%.3f", baseline$err_good), collapse = ", ")` respectively -
four different tasks and instructions give four different genuine error rates -
yet every series with a usable fast tail climbs toward chance there. The
contamination signature is task-general; the baseline is not.

The two `rtdists` datasets carry the point less cleanly than the Illusion Game,
in opposite ways. In lexical decision the fast bins hold only ten to fifteen trials
each, so they scatter widely around 0.5 rather than tracing it smoothly.
Brightness discrimination has the opposite problem: three highly practiced
laboratory subjects produce almost nothing below 0.2 s, and its genuine error
rate is already high because the task is hard, so the climb from baseline to
chance is real but compressed. Neither weakens the conclusion - they are
different failure modes of the same measurement, and both point the same way.

The **outlier component covers where the plateau is**, and its decline roughly
tracks the recovery of accuracy. That is the shape argument made visually: `g`
should be at full strength while responses are at chance, and fading as genuine
decisions take over.

The **rate is condition-dependent**, which the next section quantifies.

The observed shape also supports a flat component. Counts are flat from about
0.01 to 0.19 s and only rise as genuine responses begin:

```{r}
#| eval: false
#| code-fold: false

h <- hist(df$RT[df$RT <= 0.60], breaks = seq(0, 0.60, by = 0.01), plot = FALSE)
hist_fast <- data.frame(mid = h$mids, count = h$counts)
```

```{r}
#| label: fig-fasttail
#| fig-cap: "*Responses in 10 ms bins. Flat through the anticipation region, rising only where genuine decisions start to appear.*"
#| fig-height: 3

fast <- hist_fast[hist_fast$mid < 0.42, ]

ggplot(fast, aes(x = mid, y = count)) +
  geom_col(width = 0.0095, fill = "grey65") +
  coord_cartesian(ylim = c(0, max(fast$count[fast$mid < 0.32]) * 1.1)) +
  labs(x = "RT (s)", y = "number of trials")
```

## How Many Outliers, and How Variable?

The conditional accuracy function puts the boundary at about 0.30 s, so the
proportion of responses below it is a usable empirical estimate of `poutlier` -
and therefore a basis for priors.

```{r}
#| eval: false
#| code-fold: true

CUT <- 0.30

# Per participant, restricted to those with enough trials to estimate a rate
n_i <- tapply(df$RT, df$Participant, length)
f_i <- tapply(df$RT < CUT, df$Participant, sum)
keep <- n_i >= 100
p_i <- (f_i / n_i)[keep]

# Zeros have no logit, so cap them at half a trial before transforming
p_capped <- pmin(pmax(p_i, 0.5 / n_i[keep]), 1 - 0.5 / n_i[keep])

rates <- list(
  illusion = list(
    n_ppt = sum(keep),
    pooled = sum(f_i[keep]) / sum(n_i[keep]),
    q = quantile(p_i, c(0, .25, .5, .75, .9, .99, 1)),
    pct_zero = 100 * mean(p_i == 0),
    pct_above5 = 100 * mean(p_i > 0.05),
    mean = mean(p_i), sd = sd(p_i),
    logit_mean = mean(qlogis(p_capped)),
    logit_sd = sd(qlogis(p_capped)),
    logit_sd_nonzero = sd(qlogis(p_i[p_i > 0]))
  ),
  lexical = do.call(rbind, lapply(levels(sa$condition), function(cd) {
    s <- sa[sa$condition == cd, ]
    nn <- tapply(s$rt, s$id, length)
    ff <- tapply(s$rt < CUT, s$id, sum)
    pp <- (ff / nn)[!is.na(nn)]
    data.frame(condition = cd, pooled = sum(ff, na.rm = TRUE) / sum(nn, na.rm = TRUE),
               median = median(pp), min = min(pp), max = max(pp),
               n_zero = sum(pp == 0), n_ppt = length(pp))
  })),
  # Only three subjects, so per-participant spread is not informative here.
  # What this dataset shows instead is the instruction effect, in isolation:
  # the same three people, the same task, two response deadlines.
  brightness = do.call(rbind, lapply(levels(rr98$instruction), function(cd) {
    s <- rr98[rr98$instruction == cd, ]
    data.frame(instruction = cd, pooled = mean(s$rt < CUT), n_trials = nrow(s))
  }))
)
```

```{r}
#| echo: false

il <- rates$illusion
cat(sprintf("Illusion Game, %s participants with >= 100 trials\n",
            format(il$n_ppt, big.mark = ",")))
cat(sprintf("  pooled rate            %.4f  (%.2f%%)\n", il$pooled, 100 * il$pooled))
cat(sprintf("  per-participant mean    %.4f   sd %.4f\n", il$mean, il$sd))
cat(sprintf("  quantiles (0/25/50/75/90/99/100): %s\n",
            paste(sprintf("%.4f", il$q), collapse = "  ")))
cat(sprintf("  participants with none  %.0f%%\n", il$pct_zero))
cat(sprintf("  participants above 5%%   %.1f%%\n", il$pct_above5))
cat(sprintf("  on the logit scale: mean %.2f, sd %.2f (%.2f among non-zero)\n",
            il$logit_mean, il$logit_sd, il$logit_sd_nonzero))
knitr::kable(rates$lexical, row.names = FALSE, digits = 4,
             caption = "Lexical decision, per participant within each condition")
knitr::kable(rates$brightness, row.names = FALSE, digits = 4,
             caption = "Brightness discrimination, pooled within each instruction")
```

Four things follow, all of them practical.

**The 0.30 s cut is a property of the paradigm, not a constant.** Brightness
discrimination makes this unmissable: under the speed instruction
`r sprintf("%.1f%%", 100 * rates$brightness$pooled[rates$brightness$instruction == "speed"])`
of trials fall below 0.30 s, against
`r sprintf("%.1f%%", 100 * rates$brightness$pooled[rates$brightness$instruction == "accuracy"])`
under the accuracy instruction - the same three people doing the same task. Those
are not contamination. Their error rate at 0.25-0.30 s is
`r sprintf("%.3f", with(subset(caf, series == "RatcliffRouder1998 (all)"), err[which.min(abs(lo - 0.25))]))`,
essentially the genuine-decision baseline, so they are simply fast decisions
under a deadline. A cut calibrated on one task will over-count on a faster one,
which is exactly why the rest of this article estimates a rate rather than
applying a threshold. Treat the numbers below as priors for the paradigms that
produced them.

**The pooled rate is well under 1%.** About
`r sprintf("%.1f%%", 100 * il$pooled)` in the Illusion Game, and
`r sprintf("%.2f%%", 100 * rates$lexical$pooled[rates$lexical$condition == "accuracy"])`
to `r sprintf("%.1f%%", 100 * rates$lexical$pooled[rates$lexical$condition == "speed"])`
in lexical decision depending on instruction. On the logit scale that is around
`r sprintf("%.1f", qlogis(il$pooled))`, so a weakly informative prior such as
`normal(-5, 1)` on the `poutlier` intercept is centred at about 0.7% and puts
roughly 95% of its mass between 0.1% and 5% - tight enough to keep chains away
from the region where half the dataset is called contamination.

**It is extremely unevenly distributed.** The per-participant standard deviation
(`r sprintf("%.4f", il$sd)`) is about four times the mean
(`r sprintf("%.4f", il$mean)`), the median participant has
**`r sprintf("%.0f%%", il$pct_zero)`** none at all, and the maximum is
`r sprintf("%.1f%%", 100 * il$q[["100%"]])`. This is not a homogeneous nuisance
rate: most people never fast-guess, and a small minority do it constantly.

**So a random intercept on `poutlier` is worth it**, and the observed spread
suggests what to expect: a logit-scale between-participant SD of roughly
`r sprintf("%.1f", il$logit_sd)`, which is large. A prior like `exponential(1)`
or a half-`normal(0, 1)` on that SD is in the right range.

::: {.callout-warning}
Two caveats. The logit-scale figures are inflated by the
`r sprintf("%.0f%%", il$pct_zero)` of participants with no fast responses, whose
rate has to be capped before transforming - the distribution is a spike near zero
plus a long right tail, not a normal on the logit scale, so a random intercept
will fit it only approximately. And participants at the top of the range are not
contributing noisy data so much as not doing the task; screening them is a
separate decision from modelling the rate.
:::

## Priors Are Not Optional

::: {.callout-important}
`brms` gives a **flat, improper prior** to the intercept of any custom-family
parameter it does not recognise, which here means both `ndt` and `poutlier`.
Fitting without priors does not fail loudly - it returns numbers, and the numbers
are nonsense:

```
                            Estimate  Rhat  Bulk_ESS
ndt_Intercept      -2.137016e+14      2.43       4.9
poutlier_Intercept  3.010229e+14      2.20       5.2
```

This is not confined to awkward datasets: a clean simulation with a single
shift and 2% fast contamination fails the same way. Use `cogmod_priors()`.
:::

The likelihood has **two flat directions**, and a flat prior turns each of them
into an improper posterior.

`poutlier` toward 1. The likelihood becomes `g` alone. On a logit link that
region sits at infinity: past about `logit(poutlier) = 40` the gradient with
respect to `poutlier` is identically zero, so the chain random-walks outward
forever. It has infinite volume, and against a flat prior infinite volume wins.
Since 0.2.1 it is at least a far *worse* fit than it was - thousands of
log-likelihood units below the sensible mode rather than hundreds - because a
half-Normal cannot explain a slow response at all, so `ndt` and the decision
parameters no longer drop out of the density completely there. The prior is
still required.

`ndt` toward 0. On a log link this is the other infinity. Once `ndt` is
negligible against the fastest response the model is just an unshifted LogNormal,
and the gradient with respect to `log(ndt)` vanishes exponentially. This one has
nothing to do with the mixture - it is inherent to putting a positive shift on a
log scale - and it is the reason a prior on `poutlier` **alone is not enough**.

`ndt` is also **unbounded above**, where the mixture has removed the sharp
likelihood penalty that a hard boundary would supply. Only the prior discourages
an implausibly long non-decision time.

`cogmod_priors()` fills in every `ndt` and `poutlier` slot `brms` would
leave flat, for the model you are actually fitting:

```{r}
#| eval: false
#| code-fold: false

f <- brms::bf(RT ~ 1, sigma ~ 1, ndt ~ 1, poutlier ~ 1 + (1 | Participant),
              family = cogmod_lognormal())

brms::brm(f, data = df,
          prior = cogmod_priors(f, df),
          init = cogmod_init(f, df),
          stanvars = cogmod_stanvars(f),
          backend = "cmdstanr")
```

It reads `brms::get_prior()` rather than guessing, so it cannot produce a prior
that matches no parameter - `0 + Intercept`, interactions, group-level terms and
smooths are all handled because none of them are assumed. It touches only slots
`brms` left flat, and only for those two parameters: `mu` and `sigma` keep the
proper `student_t` intercepts `brms` already gives them. The result is passed
through `brms::validate_prior()`, so what comes back is a `brmsprior` object:
the same class returned by `brms::get_prior()` and `brms::prior()`. It prints
as the complete prior table, but can be composed with `c()`. 

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Overriding a prior, and the same set written by hand"

priors <- c(
  cogmod_priors(f, df),
  brms::prior(normal(-2, 0.3), class = "Intercept", dpar = "ndt")
)
```

The values regularize both flat directions while keeping the prior mass in
plausible regions. Slopes get `normal(0, 0.2)` and group-level SDs
`exponential(1)`: on a log or a logit link a flat slope prior is not as harmless
as it looks, and an unconstrained group-level SD lets individual participants
reach the flat regions even when the population intercept is well behaved.

`ndt` is a location in time, stated in seconds like everything else in these
families. `sigma` (an SD on the log scale) and `poutlier` (a proportion) are
unitless.

What each one means once pushed through its link:

```{r}
#| code-fold: true
#| code-summary: "Implied prior ranges on the natural scale"

softplus <- function(x) log1p(exp(x))

implied <- function(parameter, mean, sd, inv, scale) {
  v <- inv(qnorm(c(0.025, 0.5, 0.975), mean, sd))
  data.frame(parameter = parameter,
             prior = sprintf("normal(%.1f, %.1f)", mean, sd),
             `2.5%` = v[1], median = v[2], `97.5%` = v[3],
             `on the scale of` = scale, check.names = FALSE)
}

knitr::kable(rbind(
  implied("mu",       -1,   1,   exp,      "median decision time (s)"),
  implied("sigma",     0,   1,   softplus, "SD of log decision time"),
  implied("ndt",      -1.2, 0.2, exp,     "non-decision time (s)"),
  implied("poutlier", -5,   1, plogis,      "proportion of trials")
), digits = 3)
```

**`poutlier`.** `normal(-5, 1)` is centred at about 0.7%, in line with the
pooled rates found above, and keeps the degenerate region far into the tail.
`exponential(1)`
on the between-participant SD matches the spread observed earlier (a logit-scale
SD of about `r sprintf("%.1f", il$logit_sd)`) while still pooling hard toward a
common rate when a dataset gives no evidence of heterogeneity.

**`ndt`.** `normal(-1.2, 0.2)` puts the non-decision time between about 165
and 550 ms, concentrating prior mass on the usual range for keypress responses.
This is a convention informed by the RT literature rather than something
demonstrated here, and it is the prior most worth revisiting: a task with an
unusual motor requirement, or reaction times recorded from a different zero
point, will need it moved.

**`mu` and `sigma`.** Both are weakly informative and mostly there to keep warmup
sane; the table gives their reach. They assume RT in seconds, like everything
else in this family. Neither should be doing real work in a dataset of any size -
if changing them moves your conclusions, the model is under-identified for a
reason worth finding.

::: {.callout-note}
Two of these are location priors on **intercepts**, so they describe the
reference level only. `cogmod_priors()` also fills in slopes, with
`normal(0, 0.2)` on `class = "b"` for `ndt` and `poutlier` - tighter than the
intercept prior, since on a log or logit link a slope of that size already
moves the parameter across most of its plausible range.
:::

## Does the Choice of the Outlier Distribution Matter?

The hyperparameters of `g` are a regularizing choice, not a claim about how
guesses are distributed. So the test is not "are these the right values" but
"does the answer depend on them".

```{r}
#| eval: false
#| code-fold: true

sub <- df[df$Dataset == "IllusionGameValidation", ]
rt <- sub$RT

nll <- function(p, gf, free_w = TRUE) {
  mu <- p[1]; sg <- exp(p[2]); nd <- exp(p[3])
  w <- if (free_w) plogis(p[4]) else 0
  d <- ifelse(rt > nd, dlnorm(pmax(rt - nd, 1e-300), mu, sg), 0)
  v <- -sum(log((1 - w) * d + w * gf(rt)))
  if (is.finite(v)) v else 1e10
}

fitm <- function(gf, free_w = TRUE) {
  st <- c(log(median(rt) - 0.2), log(0.5), log(0.2), qlogis(0.01))
  if (!free_w) st <- st[1:3]
  o <- optim(st, nll, gf = gf, free_w = free_w,
             control = list(maxit = 8000, reltol = 1e-10))
  c(exp(o$par[3]), if (free_w) plogis(o$par[4]) else 0, -o$value)
}

sens <- do.call(rbind, lapply(names(candidates), function(nm) {
  r <- fitm(candidates[[nm]])
  data.frame(component = nm, ndt = r[1], poutlier = r[2], logLik = r[3])
}))
r0 <- fitm(candidates[[1]], free_w = FALSE)
sens <- rbind(sens, data.frame(component = "none (poutlier = 0)",
                               ndt = r0[1], poutlier = 0, logLik = r0[3]))
rownames(sens) <- NULL

fitinfo <- list(n = nrow(sub), n_ppt = length(unique(sub$Participant)),
                min_rt = min(rt))
```

```{r}
#| echo: false

s <- sens
s$ndt <- round(s$ndt, 4)
s$poutlier <- round(s$poutlier, 4)
s$logLik <- ifelse(s$logLik < -1e8, "did not fit", sprintf("%.1f", s$logLik))
knitr::kable(s, col.names = c("g", "ndt (s)", "poutlier", "logLik"))

ok <- sens[sens$logLik > -1e8, ]
cat(sprintf("ndt across all seven components: %.3f to %.3f s (%.0f ms range)\n",
            min(ok$ndt), max(ok$ndt), 1000 * diff(range(ok$ndt))))
```

Three things follow.

`ndt` is effectively invariant. Components differing by orders of magnitude in
shape, location and boundedness all land within about 10 ms of each other, which
is what a regularizing device should deliver.

`poutlier` is **not** invariant, ranging from
`r sprintf("%.3f", min(ok$poutlier))` to `r sprintf("%.3f", max(ok$poutlier))`.
It is a rate defined relative to an assumed reference distribution, not an
absolute proportion of bad trials, so it is not comparable across differently
specified models. Report the component alongside it.

Without any outlier component the model **cannot be fitted at all**. The last row
is a failure, not a poor estimate: with a fastest response of
`r sprintf("%.3f", fitinfo$min_rt)` s the likelihood has no usable region and the
optimizer never leaves its starting value.

## Recovery

Invariance says the estimate is stable, not correct.

```{r}
#| code-fold: true
#| code-summary: "Recovery by maximum likelihood, 25 replications per condition"

set.seed(7)

nll_rec <- function(p, x) {
  mu <- p[1]; sigma <- exp(p[2]); ndt <- exp(p[3]); poutlier <- plogis(p[4])
  d <- -sum(dcogmod_lognormal(x, mu, sigma, ndt, poutlier, log = TRUE))
  if (is.finite(d)) d else 1e10
}

fit_one <- function(x) {
  start <- c(log(median(x) - 0.8 * min(x)), log(0.5), log(0.8 * min(x)),
             qlogis(0.02))
  o <- optim(start, nll_rec, x = x, control = list(maxit = 8000))
  c(ndt = exp(o$par[3]), poutlier = plogis(o$par[4]))
}

rec <- do.call(rbind, lapply(c(0.12, 0.35), function(true_ndt) {
  v <- replicate(25, fit_one(
    rcogmod_lognormal(600, mu = -0.7, sigma = 0.5, ndt = true_ndt, poutlier = 0.02)
  ))
  data.frame(truth = true_ndt,
             ndt = sprintf("%.4f (%.4f)", mean(v["ndt", ]), sd(v["ndt", ])),
             poutlier = sprintf("%.4f (%.4f)", mean(v["poutlier", ]),
                                sd(v["poutlier", ])))
}))

knitr::kable(rec, row.names = FALSE,
             col.names = c("true ndt", "estimated ndt (sd)",
                           "estimated poutlier (sd)"))
```

Both recover, at a short non-decision time and a long one, with `poutlier` close
to its true 0.02.

## Per-Trial Outlier Probabilities

`poutlier` is a rate, not a classification. The mixture nonetheless assigns each
response a posterior probability of having come from the outlier component,
returned by `p_outlier()`:

```{r}
#| eval: false
#| code-fold: false

f <- brms::bf(RT ~ 1, sigma ~ 1, ndt ~ 1, poutlier ~ 1,
              family = cogmod_lognormal())

m <- brms::brm(f, data = df, 
               prior = cogmod_priors(f, df), 
               init = cogmod_init(f, df),
               stanvars = cogmod_stanvars(f),
               backend = "cmdstanr")

head(p_outlier(m))
```

That probability is the mixture *responsibility*, and it is a deterministic
function of the parameters, so its shape can be drawn without fitting anything.
The curve below uses the same expression `p_outlier()` evaluates, at illustrative
values (`mu = -0.9`, `sigma = 0.5`, `ndt = 0.25`, giving a median RT of
`r sprintf("%.2f", 0.25 + exp(-0.9))` s), for three different rates.

```{r}
#| label: fig-responsibility
#| fig-cap: "*Posterior probability that a response came from the outlier component, as a function of its RT. Dashed line marks `ndt`. Everything below it is attributed to the outlier component with certainty; the probability then collapses within about 150 ms and climbs again in the slow tail.*"
#| fig-height: 3.5
#| code-fold: true
#| code-summary: "The responsibility curve"

MU <- -0.9; SIGMA <- 0.5; NDT <- 0.25

responsibility <- function(rt, poutlier, mu = MU, sigma = SIGMA, ndt = NDT) {
  g <- 2 * dnorm(rt, 0, 0.2)                             # outlier component
  f <- ifelse(rt > ndt, dlnorm(pmax(rt - ndt, 1e-300), mu, sigma), 0)
  r <- poutlier * g / (poutlier * g + (1 - poutlier) * f)
  r[!is.finite(r)] <- 1   # both components vanish: attribute to the outlier
  r
}

resp <- expand.grid(rt = seq(0.005, 3, length.out = 1200),
                    poutlier = c(0.005, 0.02, 0.10))
resp$p <- responsibility(resp$rt, resp$poutlier)
resp$poutlier <- factor(resp$poutlier, labels = c("0.5%", "2%", "10%"))

ggplot(resp, aes(x = rt, y = p, colour = poutlier)) +
  geom_vline(xintercept = NDT, linetype = "dashed", colour = "grey40") +
  annotate("text", x = NDT, y = 1.05, label = "ndt", hjust = -0.2,
           size = 3.2, colour = "grey40") +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("#0072B2", "grey20", "#D55E00")) +
  scale_y_continuous(limits = c(0, 1.08), breaks = seq(0, 1, 0.25)) +
  labs(x = "RT (s)", y = "P(outlier | RT)", colour = "poutlier") +
  theme_minimal() +
  theme(legend.position = "top")
```

Three things are visible, and all three are properties of the mixture rather than
of any threshold.

**Below `ndt` the probability is exactly 1**, because the decision component has
no density there at all - nothing else could have produced the response. This is
the cliff from the first section, seen from the other side: the region that makes
an unmixed model unfittable is simply one the outlier component owns outright.

**The collapse is fast but not a step.** From 1 at `ndt` the curve falls to near
zero within roughly 150 ms, and where it crosses depends on the estimated rate:
at `poutlier = 0.5%` a 0.30 s response is
`r sprintf("%.2f", responsibility(0.30, 0.005))`, at 10% it is
`r sprintf("%.2f", responsibility(0.30, 0.10))`.
The same trial is graded differently depending on how much contamination the rest
of the dataset implies - which is the whole difference between estimating a rate
and applying a cutoff.

**It does not rise again in the slow tail.** Up to 0.2.0 it did, because the
half-t had heavier tails than the LogNormal, so far-out slow responses were
eventually better explained by the outlier component than by the model - and
`ndt` got pulled up behind them. That was a defect, and replacing the component
with a half-Normal is what removed it. The slow tail is now entirely the
decision family's business, which also means the model will not quietly absorb
slow contaminants for you: see the section below.

The grading is driven by evidence rather than a cutoff, and it agrees with the
conditional accuracy function without having been shown any accuracy data.

One caveat matters: a low probability does **not** mean a trial is clean, only
that the data cannot tell. A lapse that happens to produce a plausible reaction
time is statistically invisible, and the mixture handles it correctly by
accounting for its *rate* without pretending to identify it. A trimming rule
cannot do this - a cutoff removes only what looks extreme, which is both the
wrong trials and not enough of them.

## Predictions Leave the Outliers Out

For plotting effects the outlier component is a nuisance: it drags expected
values toward its own mean and adds a spike of implausibly fast draws
to any posterior predictive sample. It is also, as the sensitivity analysis
above showed, a regularizing device rather than a claim about how guesses are
distributed - so simulating from it means simulating from something the model
does not assert.

So `posterior_predict()` and `posterior_epred()` describe the 
**decision process alone** by default, as if `poutlier` were zero. Nothing needs to be done to get
this, and it carries through the packages built on `brms`, like `modelbased` or `marginaleffects`.

In the case where one wants the full mixture, for instance for posterior predictive checks,
cogmod implements the `with_outliers()` function to restore it (and `without_outliers()` to reverse it). 
The same flag can be set up front with `cogmod_lognormal(predict_outliers = TRUE)` during
model fitting.

```{r}
#| eval: false
#| code-fold: false

brms::pp_check(with_outliers(m))
```

The reason why it is implemented as a flag (toggleable with a function) that rides on the family rather than as an argument of `predict()` is because
`brms` sends the `...` of `posterior_predict()` and `posterior_epred()` to
`prepare_predictions()`, not down to the family method - `posterior_epred`
reaches it with `prep` and nothing else - and `insight`, `modelbased` and
`marginaleffects` inherit that. So `posterior_epred(m, predict_outliers = TRUE)`
is *silently ignored* rather than erroring. Carrying it on the object is what
makes it work through all of them.

`log_lik()` has no equivalent. The likelihood *is* the mixture; dropping a
component from it would not be a different summary of the same model, but a
different model. One consequence is worth knowing: `posterior_predict()` and
`log_lik()` do not describe the same distribution by default. This also desyncs
`loo_pit()`, `loo_predict()` and `bayes_R2()` from `loo()`, not just hand-rolled
checks - anything that compares a simulated replicate against the likelihood
should be run on `with_outliers(m)`.

## Slow Responses Are Your Problem

The mixture is designed to handle primarily fast contamination. Extending the same logic upward
fails: in @fig-caf the error rate *rises again* for slow responses, from
`r sprintf("%.3f", min(subset(caf, series == "Illusion Game (all)")$err))`
around 0.5 s to
`r sprintf("%.3f", with(subset(caf, series == "Illusion Game (all)"), err[which.max(lo)]))`
in the slowest bin. That is not contamination - slow trials are hard trials, so
difficulty confounds RT and accuracy at that end.

A slow contaminant is statistically indistinguishable from the right tail of the
RT distribution itself, so it cannot be identified, and leaving such trials in
biases `ndt` upward. Filter implausibly slow responses before fitting. Unlike
fast responses, the model cannot do it for you.

## Summary

- A shifted RT distribution gives zero density below `ndt`, bounding
  non-decision time by the fastest observed response. On untrimmed data this
  makes the unmixed model unfittable, not merely biased.
- Mixing in an outlier component removes the boundary, keeps the log-density
  smooth, and lets `ndt` be estimated directly in seconds.
- `g` must be flat at the origin, must have died away before real decisions
  begin, must never be exactly zero, and must possess a finite mean.
  Half-Normal(0, 0.2 s) is the only common candidate that satisfies all four;
  the half-t used up to 0.2.0 failed the second, competing with the model for
  the slow tail. The scale is a
  constant in **seconds**, so reaction times have to be supplied in seconds.
- Priors on `ndt` and `poutlier` are required, not advisory: `brms` leaves both
  flat and the likelihood has two flat directions, so the posterior is improper
  without them. `cogmod_priors()` fills them in.
- Accuracy confirms the premise independently: fast responses climb toward
  chance across three very different paradigms.
- The rate is well under 1% pooled, but wildly uneven between participants -
  most have none, a few have a great many - so a random intercept on `poutlier`
  is usually justified.
- Where the fast region ends is a property of the paradigm. Under a speed
  deadline, genuine decisions cross a cut calibrated on a slower task, which is
  why the model estimates a rate instead of applying a threshold.
- Predictions exclude the outlier component by default; `with_outliers()` puts
  it back for a like-for-like posterior predictive check.
- `ndt` moves ~10 ms across wildly different components; `poutlier` does not
  transfer between them and should be reported with its component.
- Slow contamination is unidentifiable and must be filtered beforehand.

```{r}
#| eval: false
#| echo: false

# Writes the cache read at the top of this article.
saveRDS(
  list(desc = desc, hist = hist_fast, caf = caf, baseline = baseline,
       sensitivity = sens, fitinfo = fitinfo, rates = rates),
  "../models/outliers.rds",
  compress = "xz"
)
```
