---
title: "Improving Sampling Efficiency and Performance"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
format:
  html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(cogmod)
library(brms)
library(cmdstanr)
```

Evidence accumulation models are slow and expensive to sample, and a fully mixed model
with a comprehensive random effects structure can easily take hours or days (or weeks) 
for real-life datasets. Below are some suggestions to help.

## The Model

Below is an example of a mixed DDM, with random intercepts and slopes on most distributional parameter of interest.
We also added a random intercept to `poutlier`, with the assumption that the proportion of very fast trials 
(anticipatory / inhibition failure responses vary across participants).

```{r}
#| eval: false
#| code-fold: false

f <- bf(
  RT | dec(Error) ~ Condition + (1 + Condition | Participant),
  boundary ~ Condition + (1 + Condition | Participant),
  bias ~ Condition + (1 + Condition | Participant),
  ndt ~ Condition + (1 + Condition | Participant),
  poutlier ~ 1 + (1 | Participant),
  sigmadrift = 0,
  sigmabias = 0,
  sigmandt = 0,
  family = cogmod_ddm()
)

m <- brm(f,
  data = df,
  prior = cogmod_priors(f, df),
  init = cogmod_inits(f, df),
  stanvars = cogmod_stanvars(f),
  chains = 4, iter = 2000, backend = "cmdstanr"
)
```


## Approximating with Faster Algorithms

Before committing to a long MCMC run, it is often worth getting a quick,
approximate answer first. [Pathfinder](https://mc-stan.org/docs/reference-manual/pathfinder.html)
(Zhang et al., [2022](https://arxiv.org/abs/2108.03782)) is a good first
choice and significant improvements over other Variational Inference (VI) algorithms 
like *meanfield ADVI* or *full-rank ADVI*. It is implemented in `cmdstanr` and 
can be used in `brms` with the `algorithm = "pathfinder"` argument.

```{r}
#| eval: false
#| code-fold: false

m_pathfinder <- brm(f,
  data = df,
  prior = cogmod_priors(f, df),
  init = cogmod_inits(f, df),
  stanvars = cogmod_stanvars(f),
  backend = "cmdstanr",
  algorithm = "pathfinder",
  chains = 16,
  single_path_draws = 4000,
  max_lbfgs_iters = 8000
)
```

Note that `chains` does not mean MCMC chains here, but controls the number of
 Pathfinder **paths**, and more paths give
the importance resampling step more diverse material to draw from. Increasing
`single_path_draws` and `max_lbfgs_iters` similarly buys a better
approximation at very little extra cost.
Also, consider using the `threads` argument to parallelize the pathfinder draws across 
multiple cores.

Pathfinder draws are not a substitute for MCMC posteriors, they approximate
the posterior and are not guaranteed to be well calibrated, especially for
variance components and correlations. 
However, its results can potentially be used to **tighten the priors**,
which in turn might help with MCMC convergence and sampling efficiency.

## MCMC Performance

The next step is to leverage the compounding benefits of two kinds of parallelization:

- **Chain parallelization.** Run each of the `chains` on its own core via
  `backend = "cmdstanr"` and `options(mc.cores = parallel::detectCores())` (or
  the `cores` argument to `brm()`). This is essentially free and should always
  be on.
- **Within-chain parallelization (multithreading).** `cmdstanr` can split a
  single chain's likelihood evaluation across multiple threads with
  `threads = threading(n)`, passed to `brm()` alongside
  `backend = "cmdstanr"`. This has real overhead: Stan has to partition the
  data and reduce the per-thread results back together, so it only pays off
  once the per-observation likelihood is expensive enough and/or the dataset 
  is large enough that the reduction overhead is small relative to the work 
  being split.

```{r}
#| eval: false
#| code-fold: false

m <- brm(f,
  data = df,
  prior = cogmod_priors(f, df),
  init = cogmod_inits(f, df),
  stanvars = cogmod_stanvars(f),
  backend = "cmdstanr",
  chains = 4,
  cores = 4,
  threads = threading(4)
)
```

Combining the two means a machine with, say, 16 cores can run 4 chains with 4
threads each, or fewer chains with more threads each if warmup is the
bottleneck rather than the number of independent chains needed for
convergence diagnostics.

## High-Performance Clusters (HPCs)

Chain and thread parallelization is limited by the cores on one machine. The
next step is to spread chains across many machines - typically many short
jobs on an HPC scheduler rather than one long job. Because warmup dominates
total sampling time for these models, recruiting many nodes for a **short**
run (fewer post-warmup iterations each) is usually more efficient than
recruiting a few nodes for a long one: each job pays the fixed compilation
and warmup cost once, but the post-warmup work is what actually needs to
scale with the number of draws you want.

A typical job array (SLURM shown here) runs one or two chains per node,
threading each chain across the cores left over after dividing them among the
node's chains:

```{r}
#| eval: false
#| code-fold: false

task_id <- as.numeric(Sys.getenv("SLURM_ARRAY_TASK_ID", unset = "1"))
total_cores <- as.numeric(Sys.getenv("SLURM_CPUS_PER_TASK", unset = "2"))

chains_per_node <- 2
threads_per_chain <- total_cores / chains_per_node  # e.g. 16 cores / 2 chains = 8 threads per chain

warmup <- 1000
iter <- warmup + 200  # short post-warmup draw per node; combined across nodes below

m_node <- brm(f,
  data = df,
  prior = cogmod_priors(f, df),
  init = cogmod_inits(f, df),
  stanvars = cogmod_stanvars(f),
  backend = "cmdstanr",
  chains = chains_per_node,
  cores = chains_per_node,
  threads = threading(threads_per_chain),
  warmup = warmup,
  iter = iter,
  seed = task_id,
  file = paste0("m_node_", task_id, ".rds")
)
```

Each array task produces its own `brmsfit` with a handful of chains. Once all
tasks have finished, combine them into a single fit with
`brms::combine_models()`, which concatenates the post-warmup draws across
fits (they must share the same model structure and data):

```{r}
#| eval: false
#| code-fold: false

fits <- lapply(list.files(pattern = "^m_node_.*\\.rds$"), readRDS)
m <- do.call(brms::combine_models, fits)
```

The result behaves like a single `brm()` fit with `chains_per_node * n_nodes`
chains worth of draws, obtained in roughly the time a single node's chains
would have taken.

## Future Directions

A different line of work sidesteps the repeated sampling of MCMC/VI entirely: 
**amortized inference**, where a neural network is trained (once, offline, potentially at
significant upfront cost) to map observed data directly to an approximate
posterior, so that inference on new datasets afterwards is close to
instantaneous rather than requiring a fresh MCMC run each time.
[BayesFlow](https://bayesflow.org/) is the most actively developed toolkit in
this space, and has already been applied to evidence accumulation and other
cognitive models.
