---
title: "The cogmod R Package: Easy-to-Use Bayesian Cognitive Models for Reaction Times, Choices and Subjective Ratings, with Applications to Computational Neuropsychology"
shorttitle: "The cogmod R Package"
author:
  - name: Dominique Makowski
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybendinglab.com/
    corresponding: true
    # roles:
    #   - Conceptualization
    #   - Software
    #   - Formal Analysis
    #   - Visualization
    #   - Writing – original draft
    #   - Writing – review & editing
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
      - ref: "id2"
        name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
      # - ref: "id3"  # Email to comply with the psykarix about the email on the title page
      #   name: "*D.Makowski@sussex.ac.uk*"
      #   department:  ~
blank-lines-above-author-note: 2
author-note:
  status-changes:
    affiliation-change: null
    deceased: null
  disclosures:
    study-registration: null
    data-sharing: null
    related-report: null
    conflict-of-interest: null
    financial-support: null
    gratitude: null
    authorship-agreements: null
abstract: "Cognitive tasks produce structured distributions of reaction times, choices, and subjective ratings, yet these data are often reduced to condition means before analysis. This can obscure differences in variability, distributional shape, and the latent processes that generate behaviour. We introduce `cogmod`, an R package that brings Bayesian cognitive models into the regression workflow used by psychologists. `cogmod` implements descriptive response distributions, evidence accumulation models, and models for bounded ratings as brms custom families, allowing model parameters to be predicted by covariates and random effects using familiar formula syntax. Fitted models remain standard brmsfit objects and therefore support established tools for priors, diagnostics, posterior predictive checks, model comparison, and post-processing. Across worked examples, we show that this framework enables direct comparison of competing decision models, quantifies the reliability of model parameters as individual-level measures, and reveals clinically relevant differences that are invisible in conventional mean-based analyses. By placing models spanning the descriptive-to-generative continuum within a common Bayesian regression framework, `cogmod` lowers the practical barrier to richer analyses of cognitive data and makes the choice of model an empirical question rather than a software constraint. `cogmod` is openly available at <https://github.com/DominiqueMakowski/cogmod>."
keywords:
  - cognitive modelling
  - Bayesian statistics
  - reaction times
  - evidence accumulation
  - subjective ratings
  - computational neuropsychology
  - R
impact-statement: null
word-count: false
floatsintext: true
numbered-lines: false
bibliography: bibliography.bib
suppress-title-page: false
link-citations: true
mask: false
lang: en
language:
  citation-last-author-separator: "and"
  citation-masked-author: "Masked Citation"
  citation-masked-date: "n.d."
  citation-masked-title: "Masked Title"
  email: "Email"
  title-block-author-note: "Author Note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; [credit.niso.org](https://credit.niso.org)) as follows:"
  title-impact-statement: "Impact Statement"
fontsize: 12pt
execute:
  cache: true
  echo: true
  eval: false
format:
  apaquarto-html:
    toc: true
  apaquarto-docx:
    toc: false
  apaquarto-typst:
    keep-typ: true
    list-of-figures: false
    list-of-tables: false
    toc: true
    papersize: "us-letter"
    numbersections: true
  apaquarto-pdf:
    documentmode: jou  # "man" (manuscript, single column) for submission; "jou" for a typeset look
    keep-tex: true
    pdf-standard: ua-2
editor_options:
  chunk_output_type: console
---

\clearpage

<!-- ==========================================================================
     REFERENCE PAPERS - for writing style, structure and framing
     ==========================================================================

     TARGET JOURNAL
     Behavior Research Methods (Springer). Chief editors: Erin M. Buchanan,
     Dora Matzke.  https://link.springer.com/journal/13428
     Note: Dora Matzke is a co-author of the EMC2 paper below. Frame cogmod as 
     complementary, not as a replacement.

     --------------------------------------------------------------------------
     PRIMARY TEMPLATE - closest thing to this paper, in the target journal
     --------------------------------------------------------------------------
     Stevenson, Donzallaz, Innes, Forstmann, Matzke & Heathcote (2026).
       Bayesian hierarchical cognitive modeling with the EMC2 package. BRM.
       https://doi.org/10.3758/s13428-025-02869-y
       Docs: https://ampl-psych.github.io/EMC2/
       WHAT TO BORROW: the paper is organised around a *workflow* (design ->
       Bayesian specification -> sampling -> diagnostics/evaluation ->
       inference) rather than around a list of models. Their "design phase"
       pitch - bridging standard regression and cognitive modelling via
       linear-model specifications for cognitive parameters - is very close to
       our own argument, so our differentiation has to be explicit and
       generous: they own the sampler and the accumulator architectures, we own
       the general-purpose regression interface and the descriptive-to-
       generative continuum (RTs AND ratings in one framework).

     Heathcote, Lin, Reynolds, Strickland, Gretton & Matzke (2019).
       Dynamic models of choice. BRM, 51, 961-985.
       https://doi.org/10.3758/s13428-018-1067-y
       WHAT TO BORROW: how much tutorial-level detail a BRM software paper is
       allowed to carry, and how to present a model zoo without it reading as
       a catalogue.

     --------------------------------------------------------------------------
     OWN PAPERS - for voice and for the "package paper" format
     --------------------------------------------------------------------------
     Makowski, Pham, Lau, Brammer, Lespinasse, Pham, Scholzel & Chen (2021).
       NeuroKit2: A Python toolbox for neurophysiological signal processing.
       Behavior Research Methods, 53(4), 1689-1696.
       https://doi.org/10.3758/s13428-020-01516-y
       <- own paper, same journal: the closest model for tone and length.

     Fisher & Makowski (2024). SequentialSamplingModels.jl.
       Proceedings of JuliaCon. https://doi.org/10.21105/jcon.00186

     Makowski, Ben-Shachar, Wiernik, Patil, Theriault & Ludecke (2025).
       modelbased: An R package to make the most out of your statistical
       models... JOSS, 10(109), 7969. https://doi.org/10.21105/joss.07969
     Makowski, Ben-Shachar & Ludecke (2019). bayestestR. JOSS, 4(40), 1541.
       https://doi.org/10.21105/joss.01541
     Ludecke, Ben-Shachar, Patil, Waggoner & Makowski (2021). performance.
       JOSS, 6(60), 3139. https://doi.org/10.21105/joss.03139
     Ludecke, Waggoner & Makowski (2019). insight. JOSS, 4(38), 1412.
       https://doi.org/10.21105/joss.01412
       <- JOSS format is much shorter and differently structured; use these for
          voice only, not for structure.

     Makowski, Ben-Shachar, Chen & Ludecke (2019). Indices of effect existence
       and significance in the Bayesian framework. Frontiers in Psychology, 10,
       2767. https://doi.org/10.3389/fpsyg.2019.02767
       <- own non-JOSS methods paper: useful model for how to argue a
          methodological point at length with simulations.

     --------------------------------------------------------------------------
     OTHER GOOD PACKAGE PAPERS IN NON-JOSS JOURNALS
     --------------------------------------------------------------------------
     Burkner (2017). brms. J. Stat. Software, 80(1).
       https://doi.org/10.18637/jss.v080.i01
     Burkner (2018). Advanced Bayesian multilevel modeling with brms.
       R Journal, 10(1), 395-411. https://doi.org/10.32614/RJ-2018-017
       <- the custom-family / distributional-regression framing we build on.
     Wiecki, Sofer & Frank (2013). HDDM. Front. Neuroinformatics, 7, 14.
       https://doi.org/10.3389/fninf.2013.00014
     Voss & Voss (2007). Fast-dm. BRM, 39(4), 767-775.
       https://doi.org/10.3758/BF03192967
     Vandekerckhove & Tuerlinckx (2008). DMAT primer. BRM, 40(1), 61-72.
       https://doi.org/10.3758/BRM.40.1.61
     Shinn, Lam & Murray (2020). PyDDM. eLife, 9, e56938.
       https://doi.org/10.7554/eLife.56938
     Ahn, Haines & Zhang (2017). hBayesDM. Computational Psychiatry, 1, 24-57.
       https://doi.org/10.1162/CPSY_a_00002
     Peirce et al. (2019). PsychoPy2. BRM, 51, 195-203.
       https://doi.org/10.3758/s13428-018-01193-y
     de Leeuw (2015). jsPsych. BRM, 47, 1-12.
       https://doi.org/10.3758/s13428-014-0458-y
     Mathot, Schreij & Theeuwes (2012). OpenSesame. BRM, 44, 314-324.
       https://doi.org/10.3758/s13428-011-0168-7
       <- the last three are BRM's house style for tool papers: short intro,
          heavy on worked examples, explicit about limitations.

     --------------------------------------------------------------------------
     METHODOLOGICAL FRAMING
     --------------------------------------------------------------------------
     Hedge, Powell & Sumner (2018). The reliability paradox. BRM, 50, 1166-1186.
       https://doi.org/10.3758/s13428-017-0935-1
     Burkner & Vuorre (2019). Ordinal regression models in psychology: A
       tutorial. AMPPS, 2(1), 77-101.
       https://doi.org/10.1177/2515245918823199
     Kubinec (2023). Ordered beta regression. Political Analysis, 31(4).
       https://doi.org/10.1017/pan.2022.20

     NOTE: DOIs above were assembled from search results and should be
     spot-checked before submission.
     ========================================================================== -->


# Introduction

## Traditional models can be misleading

Consider a dataset containing reaction times (RTs) from 20 participants tested under two
experimental conditions (25 trials each).
A traditional analysis approach would involve fitting a linear mixed model testing the effect of
condition, with participants entered as random intercepts:

```r
lmer(RT ~ Condition + (1 | Participant), 
     data = cogmod::badlm)
```

Because this analysis yields no significant difference ($b_{condition} = -0.001$, $t(996) = -0.08$, $p = .94$;
@fig-motivation B), one could misleadingly interpret and report it as an absence of any
difference in the RTs of the two conditions. 
Yet the two distributions are actually completely different (@fig-motivation A): 
condition A produces a lot of very fast and very slow responses, whereas condition B yields 
much more consistent RTs. But because the mean of their distribution is similar, 
the analysis does not capture any other differences. 
The problem is not the design, the data preprocessing, the random-effect structure or the post-processing, but the model itself: linear regressions estimate means and nothing else.
Simply changing the model to use another distribution, such as ex-Gaussian, drastically changes
the conclusions that can be drawn from the same data, revealing significant differences in the bulk of the distribution, its spread and its tail (@fig-motivation C).
Despite the massive benefits we can get from using more appropriate models, they are rarely used in practice, which is what the `cogmod` package aims to address.


![Simulated reaction times from two experimental conditions, available in the `cogmod` package. **(A)** Observed distributions; dashed lines mark the condition means. **(B, C)** The effect of condition as each model reports it, with 95% intervals; green marks an interval excluding zero, red one that does not. The linear mixed model has only a mean to report, and finds no effect in it. The ex-Gaussian splits the same data into a Gaussian bulk ($\mu$), its spread ($\sigma$) and an exponential tail ($\tau$), and finds an effect on all three; because its mean is $\mu + \tau$, the opposing effects on bulk and tail cancel into exactly the null of panel B.](figures/fig_motivation.png){#fig-motivation apa-twocolumn=true fig-align="center" width="100%"}

## From means to mechanisms

Cognitive tasks yield complex data, resulting from a sequence of processes that unfold over time.
Some responses come almost immediately, some follow a lapse of attention,
some reflect an unusually cautious pass over the stimulus, and some are
emitted before the evidence has really been weighed. 
The shape of the distribution, its mode, its spread and its tail, is
where the underlying processes leave their trace.
Most published analyses reduce all of it to a handful of condition means and
test for their difference. That reduction is itself a modelling
decision, and a costly one: most of what makes cognitive data distinctive is
discarded in the process.

<!-- Transforming the outcome is the usual first response to this, and it does -->
<!-- not help. Log- or inverse-transforming RTs before fitting a linear model -->
<!-- changes the quantity being tested: the model then compares means of -->
<!-- transformed values, which do not back-transform to mean RTs and can -->
<!-- reverse, create or mask effects relative to the untransformed scale -->
<!-- [@schramm2019transformations; see @liddell2018analyzing for the analogous -->
<!-- argument about ordinal data]. The underlying problem survives the -->
<!-- transformation anyway, since differences in shift, spread and tail weight -->
<!-- are still funnelled through a single location parameter. -->

The traditional summary-statistics approach involving *t*-tests, ANOVAs, linear
regressions, treats observations as Normally distributed around a
predicted mean (@fig-approaches). Two alternatives can be distinguished
from it. The **distributional** approach chooses a likelihood that matches
the shape of the outcome, such as an ex-Gaussian or shifted LogNormal for
reaction times [RTs, @matzke2009psychological; @balota2011moving], or an
ordered Beta for bounded ratings [@kubinec2023ordered], so that skew,
boundedness and dispersion are described rather than ignored. The
**computational** approach goes further and derives the likelihood from an
account of the process that generated the data. Evidence accumulation
models (EAMs, or sequential sampling models) treat response times as the
first-passage times of a noisy accumulation towards a decision boundary,
and their parameters map onto theoretically distinct components of the
decision: the quality of the evidence, how much of it is required before
committing, and the time spent encoding the stimulus and executing the
response [@ratcliff2008diffusion; @brown2008simplest].


![The three approaches applied to the same trials: correct responses from three participants in the lexical decision data of @wagenmakers2008diffusion, used again in Example 1. Observed reaction times are on the left; each row then shows what one model is made of and the distribution that follows from it, drawn back over the same data. The 141 ms difference between instruction conditions is first a single number, then a shift plus a shorter tail, then a statement about how evidence was accumulated. The ex-Gaussian and shifted Wald curves both use posterior means from models fitted to these trials in the package's *RT-only Models* vignette.](figures/fig_approaches.png){#fig-approaches apa-twocolumn=true width="100%"}

The case for moving past the first approach is well supported. Distributional
analyses recover effects that mean comparisons miss and disambiguate
effects that mean comparisons conflate [@balota2011moving; @deboeck2019overview].
Process models turn an experimental effect into a
statement about *which component of processing changed*, which is what
most cognitive-theoretical claims are about [@ratcliff2016diffusion; @evans2020evidence]. 
Model-based parameters are also more reliable than
the difference scores traditionally computed from the same data, making
them one of the more promising responses to the "reliability paradox"
[@hedge2018reliability; @haines2020learning].

The gap between what a task produces and what is taken from it is widest in
neuropsychological assessment, which rests almost entirely on cognitive
tasks. The indices extracted from them are total scores, completion times
and differences between condition means: quantities inherited for
historical rather than theoretical reasons, with no principled mapping onto
the constructs they are said to measure [@ratcliff2022neuropsychological].
Older adults, for instance, are reliably slower, but a meta-analysis of
diffusion model analyses locates the general age effect on boundary
separation and non-decision time, with drift rate differences that are
task-dependent and occasionally reversed [@theisen2021age]. A latency
difference reported as "cognitive slowing" could be more accurately framed as a statement about
caution or motor execution. Computational neuropsychology
[@steinke2020computational] proposes to keep the tasks and replace those
scores with parameters that refer to identifiable components of processing,
so that a deficit is characterized rather than merely detected.

The same holds for brain-behaviour correlations, a field burdened by
weak associations and low reliability. Brain-wide association studies based on 
basic behavioural indices (such as mean RTs) need thousands of participants before their effects replicate
[@marek2022reproducible], and the test-retest reliability of common
task-fMRI measures is typically low [@elliott2020test]. In contrast, parameters derived from cognitive models are more
reliable [@haines2020learning; @rouder2024signal] and closer to what a
neural account is about. The threshold adjustment made under time pressure
is a case in point: it is a model parameter rather than an observable, and
seems to accurately track striatal and pre-supplementary motor activation, and
cortico-striatal connection strength, across individuals
[@forstmann2008striatum; @forstmann2016sequential; @wiecki2015modelbased].
Whether the target is a clinical score or a brain-behaviour correlation,
the root cause is the same: we only get what we fit, and fitting means with linear models is a
self-imposed limitation.

## The obstacle is practical

Uptake of alternative approaches remains unfortunately modest, and the obstacle is mostly practical. 
The software landscape for fitting alternative models is fragmented.
Diffusion models are fitted in HDDM or HSSM in Python
[@wiecki2013hddm; @fengler2026hssm], in fast-dm [@voss2007fastdm], in DMAT
[@vandekerckhove2008diffusion], or in PyDDM [@shinn2020flexible]; racing
accumulator models in DMC and EMC2 in R
[@heathcote2019dynamic; @stevenson2026bayesian], or with
`SequentialSamplingModels.jl` in Julia [@fisher2024sequential]; bounded and
ordinal responses in yet another set, such as `ordbetareg`
[@kubinec2025ordbetareg]. Each
toolbox has its own syntax for specifying which
parameters depend on which conditions, its own parameterizations, defaults, and its own
output objects.

As a consequence, comparing families is difficult and therefore rare. Deciding
between an ex-Gaussian and a shifted LogNormal, or asking whether a
Weibull distribution would do as well as either, means re-fitting in different
packages under different conventions, which slows exploratory work and
leaves the choice of likelihood to be inherited from disciplinary
convention rather than tested against the data at hand. 
Second, these toolboxes often sit apart from the regression workflow
psychologists already use, such as `brms` in R [@burkner2017brms], which means missing on the
expressive power of that workflow. Every design has
to be restated in a toolbox-specific syntax, and features like complex random effects specifications, smooth terms, monotonic effects for ordinal predictors, or post-processing tools, such as marginal effects and means, are effectively out of reach [@wood2017generalized; @burkner2018advanced].

`cogmod` aims to address these limitations by facilitating the application of 
these models inside a framework researchers already use. 
Every model is defined as a `brms` custom family
[@burkner2017brms] specified with ordinary `brms` formula syntax. Everything a `brms` user
already knows then applies unchanged to a Drift Diffusion Model or an
ordered-Beta rating model: distributional regression on every parameter,
multilevel structure, priors, convergence diagnostics, posterior predictive
checks, leave-one-out cross-validation [@vehtari2017practical]. The
`easystats` ecosystem [@ludecke2019insight; @makowski2019bayestestr; @ludecke2021performance] has
been adapted to accommodate these families, providing first-rate post-processing tools.

We describe below the design of the package and the models it provides,
then give three worked examples, one per family of models, to illustrate the workflow 
and applications. 

# Design and Implementation

## Models as `brms` custom families

`brms` translates an R formula into Stan code. Its *custom family*
mechanism allows users to supply their own likelihood: a `customfamily`
object declaring the distributional parameters and their link functions,
plus a `stanvars` object injecting the Stan functions that evaluate the
log-density. `cogmod` packages both for each model, and additionally
provides the post-processing methods (`log_lik_*`, `posterior_predict_*`
and `posterior_epred_*`) that `brms` requires in order to compute
information criteria and generate predictions used in post-processing.

In practice, using a `cogmod` model requires two arguments beyond a
standard `brm()` call: `family` and `stanvars`. Two further helpers,
`cogmod_priors()` and `cogmod_inits()`, are strictly speaking optional but
should be treated as part of the call: they provide adapted chain-initialization 
values and weakly informative priors on the sensitive parameters to limit
convergence issues and other sampling pathologies.

```{r}
#| eval: false

library(cogmod)
library(brms)

# Specify the formula using brms' bf()
f <- bf(
  RT ~ Condition +
    (Condition | Participant),
  sigma ~ Condition +
    (Condition | Participant),
  ndt ~ Condition +
    (Condition | Participant),
  family = cogmod_lognormal()
)

# Fit the model
m <- brm(
  f,
  data = df,
  prior = cogmod_priors(f, df),
  init = cogmod_inits(f, df),
  stanvars = cogmod_stanvars(f),
  backend = "cmdstanr"
)
```

Everything else is `brms`. The formula syntax is the usual one, so any
parameter can be predicted by covariates, given random effects, splines,
autocorrelation structures or monotonic effects. The fitted object is a
plain `brmsfit`, so `summary()`, `pp_check()`, `loo()`, `emmeans`,
`bayesplot` and the `easystats` functions work out of the box.
Contrary to many alternative toolboxes, here each distributional parameter is a regression target with its own formula.
Testing the effect of a manipulation on various parameters becomes thus trivial, 
as is giving each participant their own parameters through random effects. 
This allows the location-scale modelling advocated by @williams2021beneath, in which the
dispersion of a participant's RTs is a person-specific quantity with its
own predictors, to be the default rather than a special case.


## Relation to existing software

@fig-overview shows the families available at the time of writing.
As adding a new model is a self-contained
piece of work, the set is expected to grow to track cutting-edge advances in computational psychology.
The package's documentation website contains the list of available models, with each family's distributional parameters, link
functions and defaults, as well as hands-on tutorials.

![Response formats covered by `cogmod` and the main families available for each at the time of writing; the package website carries the current list. Six further descriptive RT families are also available (`cogmod_weibull()`, `cogmod_logweibull()`, `cogmod_invweibull()`, `cogmod_gamma()`, `cogmod_invgamma()` and `cogmod_bisa()`), and the package's *RT-only Models* vignette compares them all on the same data. Every family comes with a `_stanvars()` function supplying the corresponding Stan code, and with `d*()` and `r*()` functions for density evaluation and simulation.](figures/fig_overview.png){#fig-overview apa-twocolumn=true width="100%"}



`cogmod` is not a replacement for the specialized evidence accumulation
toolboxes, and the division of labour is worth stating. EMC2
[@stevenson2026bayesian] and HSSM [@fengler2026hssm] provide sampling
algorithms tailored to the geometry of accumulator models, hierarchical
structures designed for them and, in HSSM's case, likelihood
approximation networks that make otherwise intractable variants
estimable. For a study whose scientific question *is* the architecture
of the decision process, those are the right tools, and they will be
faster and more capable at the far end of model complexity.

`cogmod` targets a different (and arguably more common) situation: a researcher
with an experimental dataset, a regression-shaped question and a need
for a likelihood that respects the data. For that user the relevant
comparison is between a linear model and something better, not between
one accumulator toolbox and another. Moving from `brm(RT ~ Condition)`
to a shifted LogNormal or a Wald model means
changing two arguments rather than changing software, and at every step
the analysis remains a regression, so most of the analyst's expertise and skills
continue to apply.

# Example 1: Many Decision Models, One Workflow

To illustrate the typical workflow of fitting evidence accumulation models, 
we will fit them to the lexical decision task data of @wagenmakers2008diffusion (Experiment 1),
distributed in the `rtdists` package as `speed_acc`: participants classified
letter strings as words or non-words under instructions emphasising either
**speed** or **accuracy**. Responses slower than 2 s were excluded. This
example uses three participants ($n = 4{,}620$ trials, 7.3% of errors),
pooled without random effects for clarity and simplicity.

In order to fit a basic Drift Diffusion Model (DDM) in `cogmod`, one needs to specify 
1) a formula for each of the parameters, 2) the model family (`cogmod_ddm()`) and attach the custom family Stan code with `cogmod_stanvars()`.
As it is common in the field, the "decision" being modelled
is **correct vs. error**, which is added to the formula with a `dec()` term
on the response side of the formula. Everything else is the ordinary
`brms` syntax:

```{r}
#| eval: false
#| echo: false

data(speed_acc, package = "rtdists")
df <- data.frame(
  Condition = unname(c(accuracy = "Accuracy", speed = "Speed")[
    as.character(speed_acc$condition)]),
  RT = speed_acc$rt,
  Error = as.integer(as.character(speed_acc$response) !=
                       as.character(speed_acc$stim_cat)),
  Participant = as.integer(as.character(speed_acc$id))
)
df <- df[df$Participant %in% c(1, 2, 3) & df$RT <= 2, ]
```

```{r}
#| eval: false

f <- bf(
  RT | dec(Error) ~ Condition,
  boundary ~ Condition,
  bias ~ 1,
  ndt ~ 1,
  sigmadrift = 0, 
  sigmabias = 0, 
  sigmandt = 0,
  family = cogmod_ddm()
)

m_ddm <- brm(f, data = df,
             prior = cogmod_priors(f, df),
             init = cogmod_inits(f, df),
             stanvars = cogmod_stanvars(f)) |>
  add_criterion("loo")
```

The first unnamed parameter, referred to as `mu` in the Stan code as per `brms` convention, 
actually corresponds to the drift rate.
The other parameters correspond to boundary separation (`boundary`), starting
point bias (`bias`), and non-decision time (`ndt`), respectively.
Estimating non-decision time is a common source of difficulty in fitting
diffusion models: a shifted distribution assigns zero density to any response
faster than `ndt`, which puts a hard wall in the likelihood at the fastest
observed response. In `cogmod`, every family that carries an `ndt` therefore mixes in a
small contaminant component, of weight `poutlier`, that keeps early responses
at positive density. The wall becomes a finite cost, and `ndt` is estimated
directly in seconds rather than relative to a sample statistic. 

The three `sigma*` arguments fix the between-trial variabilities of the
full seven-parameter diffusion model [@ratcliff2008diffusion] at zero,
giving the "simple" four-parameter DDM, which is a common simplification, since
the variability parameters are expensive to sample and notoriously hard
to recover [@boehm2018estimating]. The
effect of `Condition` is also estimated for the boundary separation,
which is the textbook hypothesis for a speed-accuracy manipulation, and the
results support it: boundary separation drops sharply under speed
instructions (95% CI on the linear predictor $[-0.65, -0.55]$).

Whether that is the right model, however, is a separate question, and answering it
means fitting the alternatives, which only involves
a few code changes: a different `family`, its `stanvars`, and whichever
distributional parameters that family owns. The **DDM-5** is the model
above with between-trial variability in drift rate freed
(`sigmadrift ~ 1`). The **LogNormal race** [LNR, @rouder2015lognormal],
the **linear ballistic accumulator** [LBA, @brown2008simplest] and the
**racing diffusion model** [RDM, @tillman2020sequential] replace the
single diffusion with two
accumulators racing towards a threshold, and differ in how each
accumulator gets there: by drawing its finishing time from a LogNormal, by
accumulating linearly with between-trial variability in its rate, or by
diffusing with within-trial noise alone. Because each fit is a `brmsfit` carrying a `loo` criterion,
comparing the five is one call, and @fig-eam B puts the result of that call
against what each architecture cost to sample (the ranking itself is three
participants' worth of evidence; it is an illustration of the workflow, not a
result):

```{r}
#| eval: false

loo_compare(m_ddm, m_ddm5, m_lnr,
            m_lba, m_rdm)
```

![Five evidence accumulation models fitted to the choices and reaction times of three participants. **(A)** Posterior predictive check: observed trials as histograms, 100 predictive draws per model as faint lines and their pooled density in bold, with correct responses drawn upwards and errors downwards. Both halves are scaled by the *proportion* of that response, so the area under each curve is its predicted frequency and the check is on the joint distribution of choices and RTs. Draws are taken with `with_outliers()`, so the check is on the same mixture the likelihood scores. Predicted RTs above 3 s (0.05% of draws, produced by near-zero drift rates in the races) are excluded. **(B)** Predictive accuracy against sampling cost. Both panels come out of the ordinary `brmsfit` methods, and neither needed the models to be refitted.](figures/fig_eam.png){#fig-eam apa-twocolumn=true width="100%"}


# Example 2: Parameter Variability

Beside decision-making models, `cogmod` also ships a large collection of 
RT-only models, spanning the descriptive-to-generative continuum: the ex-Gaussian and the
shifted LogNormal, well established in the RT literature, the shifted Wald, the single-accumulator LBA and its recinormal (LATER model) special case whose parameters refer to an accumulation process, and a longer
tail of families - log-Student, log-Gamma, Birnbaum-Saunders, and several
Weibull and Gamma variants. While they can be easily compared with the same workflow as above, 
the question this example asks instead concerns interindividual variability.

Consider a simple detection task typical of attentional paradigms: a stimulus appears, the participant
responds as fast as they can, with 80 times per three experimental conditions A, B, and C. 
RT data of thirty people are modelled with a shifted Log-Normal mixed model, with every
parameter free to vary by condition and by participant (random slopes):

```{r}
#| eval: false

f <- bf(
  RT ~ Condition +
    (Condition | Participant),
  sigma ~ Condition +
    (Condition | Participant),
  ndt ~ Condition +
    (Condition | Participant),
  family = cogmod_lognormal()
)
```

Read as an experiment, the result seems clear and straightforward. Condition B
significantly slows responses relative to A on the location parameter
(B-A: $b_\mu = 0.10$, 95% CI $[0.04, 0.17]$), but condition C is not different
(C-A: $b_\mu = -0.02$, 95% CI $[-0.15, 0.10]$). Neither condition impacts the
dispersion or the non-decision time. The temptation is to treat the one significant effect as
the finding, i.e., the quantity the task "measures" and the natural candidate for a
score to correlate with a questionnaire, a phenotype or a scan. It is in fact
the *worst* of the six for that purpose, and condition C, the null result, is
among the best.

The reason is that a fixed effect and an individual difference are answers to
different questions, and a design powered for the first says nothing about the
second [@hedge2018reliability; @haines2020learning; @rouder2024signal; @makowski2023clinical]. The
distinction is decisive wherever the point of a task is to produce a number
that characterises a *person* rather than a population - which is to say
throughout neuropsychology and differential research, where tasks are
routinely adopted on the strength of a group effect alone. The
random-effect SDs already hint at it: 0.01 for the condition B slope against
0.28 for the condition C slope, more than twenty times larger. Condition B
slows *everyone* by about the same amount, which is exactly what makes its
group effect so precise.
Condition C slows some participants dramatically and speeds others up just as
much, which averages to nothing.

Participant-level estimates can be extracted with
`modelbased::estimate_grouplevel()`, and their usability as individual
scores can be summarised by the **Variance-Over-Uncertainty Ratio**
(D-vour), implemented in `performance::performance_dvour()`
[@ludecke2021performance]:

$$
D_{\text{vour}} = \frac{\sigma_{\text{between}}^2}{\sigma_{\text{between}}^2 + \mu_{\text{within}}^2}
$$

where $\sigma_{\text{between}}^2$ is the observed variability of the
participant-level estimates and $\mu_{\text{within}}^2$ their mean
squared uncertainty. It is bounded in $[0, 1]$, and values near 1 mean
participants differ far more than we are uncertain about where each of
them stands. It is close to an intraclass correlation, but its
denominator uses the *uncertainty of the point-estimates* rather than the raw
within-participant variance, so that, unlike an ICC, collecting more
trials per participant increases it; and relates to the signal-to-noise ratio advocated by @rouder2024signal.

@fig-dvour shows the result for every parameter of the model. 
The intercept (the location parameter of Condition A) and condition C
both have a broad marginal band and points that fan out well past their
intervals, which means people are well separated on these parameters. 
In contrast, the effect of Condition B vs. A (the only significant fixed effect),
has points around zero inside intervals
that dwarf their spread.

![Individual-level estimates for every parameter of the mixed shifted LogNormal. Each panel is one parameter: points are the 30 participants' deviations from the group, ordered by estimate, with 95% intervals; the shaded band above is the marginal distribution of those point estimates. A parameter is usable as an individual score when the points spread further than the intervals are wide - which is what D-vour, printed in each panel, measures. Green marks values above 0.5. `estimate_grouplevel()` labels the `mu` block "conditional", since it is the response formula; it is written as `mu` here to match the family's parameter names.](figures/fig_dvour.png){#fig-dvour apa-twocolumn=true width="100%"}

While this simulated data was created specifically with no inter-individual variability
in the `sigma` and `ndt` parameters, real data often look
different: RT variability can often discriminate between individuals better
than the mean does [@williams2021beneath], and non-decision time is a
natural candidate for a stable person-specific quantity. But in order to be able
to assess that, one needs to fit appropriate models. The recommendation is simple:
to model and quantify variability for every parameter, and report it alongside the "main" effects.

# Example 3: Illustration in a Clinical Population

The data so far have been a laboratory experiment and a simulation. The
case for computational neuropsychology, though, is illustrated with clinical
data, where trials are few, samples are heterogeneous and where more appropriate
models can significantly improve insights gained from the data.

We showcase this on the UCLA Consortium for Neuropsychiatric Phenomics dataset
[@poldrack2016phenome], openly available on OpenNeuro as `ds000030`, and
take the task-switching data from 42 participants diagnosed with
ADHD and 125 healthy controls. Participants
classified a coloured shape by either its colour or its shape, with the
relevant dimension cued on each trial and switching on a quarter of
them. Keeping correct responses between 200 ms and 2 s leaves 14,543
trials, at most 96 per participant.

The standard index for this task is the *switch cost*, the difference
between a participant's mean RT on switch and non-switch trials, read as
a measure of executive control. A *t*-test shows no significant difference between the groups: 
Controls average 111 ms and the ADHD group 113 ms, $t(82) = -0.13$, $p = .90$. A linear mixed
model agrees: a large switch cost overall ($b = 0.11$ s, $t = 13.3$,
$p < .001$, so the manipulation worked), no group difference in it
($p = .89$), and no reliable group difference in overall speed
($p = .12$). Taken at face value, this task detects no deficit in this sample.

We re-analyzed this dataset using a shifted Log-Normal model in which diagnosis and switching predict
the location, the dispersion and the non-decision time, with random
effects by participant:

```r
f <- bf(
  RT ~ Diagnosis * Switching +
    (Switching | Participant),
  sigma ~ Diagnosis * Switching +
    (1 | Participant),
  ndt ~ Diagnosis + (1 | Participant),
  family = cogmod_lognormal()
)
```

The model agrees about the switch cost: the interaction is null on the
location and on the dispersion. But diagnosis registers on all three parameters at once, and in opposing
directions. On the response scale (@fig-adhd), the ADHD group's location parameter 
is 96 ms slower (95% CI $[26, 178]$),
their non-decision time is 48 ms faster (95% CI $[-75, -16]$), 
and their log-scale dispersion is lower (95% CI $[-64.9, 4.4]$).
The two effects largely cancel, leaving the total mean
difference non-significant (95% CI $[-17, 124]$), which is what the null the *t*-test reported. 

![The same task-switching data under two analyses (42 ADHD, 125 controls). (A) The conventional switch-cost index, computed for each participant as their mean RT on switch trials minus their mean RT on non-switch trials. The distributions of these individual switch costs do not separate the groups. Dashed lines mark the group means. (B) Posterior distributions of the ADHD - control differences under the shifted LogNormal model, on the response scale. The total mean RT difference is close to zero because a slower decision component is accompanied by a faster non-decision component.](figures/fig_adhd.png){#fig-adhd apa-twocolumn=true width="100%"}

This is the cancellation of @fig-motivation, present in real clinical data: 
two components move in opposite directions, the mean
records their sum, and a difference score computed from means is
therefore blind to both. No amount of statistical power on the switch
cost would have recovered this. 

Note however that the decision and non-decision
parameters trade off in the posterior ($r = -0.44$), so the split
between them is less certain than either interval taken alone; what is
robust is that a group difference exists in the shape of the
distributions and not in their means. And this is one task in one
sample, analysed post hoc: it is a demonstration that the question
becomes askable, not a claim about the neuropsychology of ADHD.

Examples 2 and 3 are two halves of one clinical argument. A group
difference in a parameter, as here, is a statement about a population;
using that parameter as a patient's score requires the check of Example 2,
run on the same fit. Both come out of one model, and neither requires the
analyst to leave the regression model they started from.

# Example 4: Subjective Ratings

<!-- NOTE FOR FUTURE EDITS: the CHOCO (Choice-Confidence) family is
 deliberately NOT *discussed* in this paper - it is the subject of a
 separate paper of its own. Do not re-introduce it in the prose, in the
 abstract, or in any caption.

 It does appear in @fig-overview, and that is intentional: the figure is
 an inventory of what the package ships, and omitting a shipped family
 from an inventory would be misleading. See the matching note in
 man/figures/make_fig_overview.py, which owns that figure. Inventory
 yes, prose no. -->

Cognitive modelling is usually applied to RTs, but the principles apply to
subjective ratings as well, which exhibit particular distribution shapes worth
characterizing. Analog (slider) scales are bounded, and responses pile up
at the endpoints, because answering "completely agree" is a different
kind of act from answering "agree quite a lot". A mean rating cannot tell
the two apart.

Consider a taste test. Participants rate how much they enjoyed each of two
dishes on a slider running from "hated it" to "loved it". One dish is a
margherita pizza: few people have strong feelings about it^[Data from the author were not included, as they would have
biased the margherita mean towards extreme enjoyment.], and the ratings pile up
in the middle. The other is a coriander salad, which a substantial minority
of people experience as tasting of soap.
Half the room loves that dish and half hates it, and almost nobody is
indifferent.

Similarly to the examples above, a *t*-test duly reports that they are equally liked ($t(296) = 0$,
$p = 1.00$), and a linear regression agrees, putting the difference at
$0.00$ with a 95% CI of $[-0.05, 0.05]$. 
`cogmod_betagate()` is `cogmod`'s reparameterization of the ordered Beta model
[@kubinec2023ordered], in which an observed 0 or 1 arises when an
underlying continuous tendency passes a lower or upper "gate". Its four
parameters are separately interpretable: `mu` is that underlying mean tendency,
`phi` the rating's consistency (the spread), `pex` the overall
probability of an extreme response, and `bex` the bias in the direction of the
extremity. As in every other family, each takes its own formula:

```{r}
#| eval: false

f <- bf(score ~ Dish,
        phi ~ Dish,
        pex ~ Dish,
        bex ~ Dish,
        family = cogmod_betagate())

m <- brm(f, data = df, 
         stanvars = cogmod_stanvars(f))
```

The four parameters separate what the mean confounded (@fig-ratings C).
Consistency collapses, from 11.2 to 0.5, so the coriander ratings are not
merely more variable but *anti*-clustered ($\phi < 1$): mass at both ends rather than
in the middle. Extreme responding (zeros and ones) rises from 3% to 41% of trials. And the
two parameters that should not move do not: the underlying mean tendency is
unchanged ($+0.02$, $[-0.03, 0.08]$), and so is the direction of the extremes
($-0.03$, $[-0.39, 0.31]$), since the coriander extremes split
evenly between "hated it" and "loved it" rather than favouring one end.

![The same 500 ratings under two models. **(A)** Observed ratings for the two dishes; the dashed lines mark the condition means, which coincide. **(B, C)** The effect of dish as each model reports it, with 95% intervals; green marks an interval excluding zero, red one that does not. The Gaussian model has only a mean to report and finds nothing in it. Beta-Gate separates how highly the dish is rated from how *consistently*, and from how often raters go to an endpoint - and it correctly reports no change in the first or in which endpoint they choose.](figures/fig_ratings.png){#fig-ratings apa-twocolumn=true fig-align="center" width="100%"}

This example generalises past taste tests. Consensus and polarisation are
routinely confounded in exactly this way whenever attitudes, symptom
severity or confidence are summarised by an average. Beta-Gate is one of 
several ways to give a bounded response its point
masses, and the zero-one inflated Beta (ZOIB) and the extended-support Beta
[@kosmidis2024extended] are others, and each attributes the endpoints to
a different mechanism. Discrete rating scales are covered too:
`cogmod_betadiscrete()` applies the same logic to Likert-type items, where
the response is one of a handful of ordered categories rather than a point
on a continuum. A response format that ordinarily calls for its own specialised package 
is here one more `family` argument, carrying the same formula syntax, the same
diagnostics and the same post-processing as any other `brms` models.

# Discussion


`cogmod` implements Bayesian cognitive models as `brms` custom families:
descriptive RT distributions, evidence accumulation models, and
models for bounded and Likert-type ratings. The design commitment is
minimal. The package supplies likelihoods and the post-processing
methods to make them work within the `brms` ecosystem, and
inherits the rest. 

The examples thread around one important *fil rouge*: *what you fit is what you get*. 
Choosing a given model is not only a claim about the shape of the data; it fixes the
vocabulary in which results can be stated. While a Gaussian model can
report that two conditions' mean differ by 141 ms, an ex-Gaussian can attribute
that difference to the bulk or to the tail, but a diffusion model will provide answers
in terms of evidence, threshold or encoding.

This point is particularly relevant for neuropsychological assessments. A
neuropsychological test is a cognitive task, so everything above applies
to it directly, with higher stakes attached. The quantity a
clinician reports, e.g., a completion time, a total score, an interference
effect, is exactly the aggregate that the opening example shows to be several
things at once, and a patient's deviation from a normative sample can 
therefore be difficult to interpret.
Model parameters turn "impaired on the Stroop" into a claim about which
component is impaired, and two patients with the same score but opposite
parameter profiles stop being indistinguishable
[@ratcliff2022neuropsychological; @steinke2020computational]. Ageing
shows what this changes in practice [@theisen2021age], and the clinical
reanalysis of Example 3 shows the sharper version of the same point: a
conventional index can report no group difference at all while two
components underneath it differ substantially. 

The reliability check that precedes it supplies the rest of the
argument. A parameter is usable as a clinical score only if it separates people by more than it is
uncertain about each of them, and this framework makes that checkable
for every parameter rather than assumed for whichever one the test
happens to report. The same approach helps on the neural side. Correlating brain
measures with mean RT or accuracy asks the brain to explain a quantity
that mixes mechanisms; correlating them with drift rate, threshold or
non-decision time asks it to explain one at a time, which is why
parameters have proven the more tractable targets in model-based
cognitive neuroscience [@forstmann2016sequential; @wiecki2015modelbased].

Advanced computational modelling of Human behaviour remained for a long 
time a relatively niche field of research, difficult to access, adopt, and apply. 
`cogmod` aims to make these tools accessible to a wider audience, to improve methodological practices with positive benefits for (neuro)psychological science.

# Limitations

A few limitations and caveats have to be noted, the first concerning performance.
General-purpose HMC on custom Stan code is
slower than a sampler built for a specific model class. The evidence accumulation 
models are particularly slow, and hierarchical
versions cost substantially more. These costs can be mitigated in several ways. 
Stan's within-chain parallelisation and distributed workflows can be exploited 
on high-performance computing clusters. Alternative Bayesian inference approaches 
such as variational inference can trade some of the guarantees or generality of full HMC 
for substantially lower computational cost.

Second, the flexibility of the `brms` interface makes it easy to
write down models the data cannot support. Regressing every parameter of
an LBA on every condition is one line of code that might appear appealing but is usually a bad idea:
evidence accumulation parameters trade off against each other, and
between-trial variability parameters are notoriously poorly identified
[@boehm2018estimating; @evans2020evidence]. The package does not, and
cannot, prevent this. Prior predictive checks, parameter recovery on
simulated data and the standard Bayesian workflow [@gelman2020bayesian]
remain the user's responsibility.

The last point is a technical boundary on what `cogmod` can become, and it follows directly
from the design. A `brms` custom family is a Stan model, so every family
shipped here needs a likelihood that can be written down in closed form
and evaluated cheaply enough for HMC. Several of the models a user might
reasonably want next do not have one: the leaky competing accumulator
[@usher2001time], diffusion processes with collapsing boundaries or
time-varying drift [@hawkins2015revisiting], and richer racing
architectures are defined by a generative simulation rather than by a
density, and where a density does exist it often requires numerical
integration, which gradient-based sampling tolerates poorly. The general
remedy is to stop requiring the likelihood. Simulation-based inference
[@cranmer2020frontier] trains a neural network on data simulated from
the model, either to approximate the likelihood [@fengler2021likelihood; @fengler2026hssm],
or to amortize the posterior directly, as in BayesFlow
[@radev2020bayesflow; @radev2023bayesflow]. That literature is moving
quickly and `cogmod` follows it closely; whether such approximations can
be exposed behind the same formula interface, so that a
simulation-defined model is specified, sampled and post-processed like
any other family, is an open question and the direction in which we hope
to extend the package.

# Conclusion

The gap between what cognitive data contain and what is extracted from
them is large, and it is sustained by friction rather than disagreement. 
Few researchers would defend a mean difference as an
adequate summary of a reaction time distribution, yet many report one,
because the alternative often involves learning a new complex set of tools. 
By implementing cognitive models as families within
a regression framework psychologists already use, `cogmod` aims to make
the better analysis the easier one; by shipping various descriptive and
generative models under one interface, it aims to make the choice
between them an empirical question rather than an inherited convention.

# Open Practices Statement

`cogmod` is open source and available at
[https://github.com/DominiqueMakowski/cogmod](https://github.com/DominiqueMakowski/cogmod), with documentation at
[https://dominiquemakowski.github.io/cogmod/](https://dominiquemakowski.github.io/cogmod/).

# Declarations

## Funding

Not applicable.

## Conflicts of interest/Competing interests

The authors have no relevant financial or non-financial interests to disclose.

## Ethics approval

Not applicable. Example 3 re-analyses openly available, de-identified, previously published third-party data [@poldrack2016phenome]; no new data involving human participants were collected for this manuscript.

## Consent to participate

Not applicable.

## Consent for publication

Not applicable.

## Availability of data and materials

The clinical dataset re-analysed in Example 3 is openly available on OpenNeuro as `ds000030` [@poldrack2016phenome]. All other examples use simulated data generated by code included in the package repository.

## Code availability

All code, including the `cogmod` package source and the scripts used to generate the examples and figures in this manuscript, is openly available at
[https://github.com/DominiqueMakowski/cogmod](https://github.com/DominiqueMakowski/cogmod).

# References

::: {#refs}
:::
